---
title: "Fold_changes_to_rpkm-tpm"
author: "Rodrigo Arcoverde Cerveira & Matt Hinchcliff"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output: html_document
abstract: Örebro study RNA-seq data analysis.
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/lab_book/")})
editor_options: 
  chunk_output_type: inline


---


```{r global.options, include=FALSE}
#setting up figures and chunks messages

knitr::opts_knit$set(echo = TRUE,
                     root.dir = getwd(),
                     fig.width = 6, fig.height = 5,
                     fig.align = "center",
                     out.width = 768,
                     fig.pos = "H",
                     warning = FALSE, 
                     message = FALSE)
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.width = 6, fig.height = 5,
                     fig.align = "center",
                     out.width = 768,
                     fig.pos = "H")

result.dir <- paste0("results/",Sys.Date(),"/")
figures.dir <- paste0("results/",Sys.Date(),"/", "figures/")
## creates result.dir with date in if not existent
ifelse(isFALSE(dir.exists(paste0("../",result.dir))), dir.create(paste0("../",result.dir),recursive = TRUE),"Result directory for today exists already!")
ifelse(isFALSE(dir.exists(paste0("../",figures.dir))), dir.create(paste0("../",figures.dir),recursive = TRUE),"Result directory for today exists already!")

options(stringsAsFactors = FALSE) 


```

## loading libraries



```{r, message=FALSE}

library(dplyr)
library(RColorBrewer)
library(EnsDb.Hsapiens.v86)
library(ggplot2)
library(clusterProfiler)
library(DESeq2)
library(caret)
library(janitor)
library(edgeR)
library(tidyr)

```

## Load and wrangle raw counts

```{r}

# Load raw counts
counts_raw <- data.table::fread("../data/processed_data_nfcore/batch_1-2/salmon/salmon.merged.gene_counts.tsv",
                                  header = TRUE, sep = "\t") %>% 
  as.data.frame()

colnames(counts_raw) <- sub("_L001|_L002", "",colnames(counts_raw))

# Identify duplicated gene_names
duplicated_genes <- counts_raw$gene_name[duplicated(counts_raw$gene_name)]

# Replace duplicated gene_names with corresponding gene_ids
counts_raw$gene_name <- ifelse(counts_raw$gene_name %in% duplicated_genes,
                               counts_raw$gene_id,
                               counts_raw$gene_name)



# Change row names to gene names
rownames(counts_raw) <- counts_raw$gene_name
counts_raw <- dplyr::select(counts_raw, -c(gene_name, gene_id))

###

# counts_raw <- data.table::fread("../data/processed_data_nfcore/batch_1-2/salmon/salmon.merged.gene_counts.tsv",
#                                   header = TRUE, sep = "\t") %>% 
#   as.data.frame() %>%
#   select(-gene_id)
# 
# 
# colnames(counts_raw) <- sub("_L001|_L002", "",colnames(counts_raw))
# # Aggregate and sum up genes with same gene symbol (basically non-coding RNAs or ribosomic)
# counts_raw <- aggregate(counts_raw[, -1], 
#                         list(gene_name = counts_raw[, 1]), 
#                         FUN = sum)
# 
# # Change row names to gene names
# rownames(counts_raw) <- counts_raw$gene_name
# counts_raw <- dplyr::select(counts_raw, -c(gene_name))

# Read antibody data
antibody_data <- read.csv("../data/antibody_data/2022-04-26_antibody-binding-avidity.csv")

# Load BTM modules
gmt <- read.gmt("../data/blood_transcription_module/BTM_for_GSEA_20131008.gmt")


```

## Load metadata and match with raw counts

```{r, results='hide'}
# Load sample metadata
sampleTable <- read.table("../data/metadata/sample_metadata_3.csv", 
                          sep = ",", header = TRUE) %>%
  dplyr::select(-c(age, sex))

# Load patient metadata
patient_metadata <- read.table("../data/metadata/patient_metadata.csv", 
                               sep = ",", header = TRUE)

# Join sample and patient metadata
sampleTable <- sampleTable %>% left_join(patient_metadata, by = "subject_ID")

# Change visits to dose and time point...
sampleTable <- sampleTable %>%
  mutate(
    dose = as.factor(plyr::mapvalues(visit,
      from = c(1, 2, 4, 5, 10, 11, 12),
      to   = c("0_dose1", "24_dose1", "0_dose2", "24_dose2", "0_dose3", "24_dose3", "48_dose3")
    )),
    group = as.factor(group)
  ) %>%
  # ...(sex) 0 and 1 to male and female...
  mutate(
    sex = as.factor(plyr::mapvalues(sex,
      from = c(0, 1),
      to   = c("male", "female")
    )),
    group = as.factor(sex)
  ) %>%
  # ...(prior covid) 0 and 1 to neg and pos
  mutate(
    prior_covid19 = as.factor(plyr::mapvalues(prior_covid19,
      from = c(0, 1),
      to   = c("neg", "pos")
    )),
    group = as.factor(prior_covid19)
  )

# Remove 24dose_2 from subject 21 (missing 0dose_2)
sampleTable <- subset(sampleTable, sample_ID != "P22761_1037_S37")

# Remove one sample from subject 7 (sample replicate)
sampleTable <- subset(sampleTable, sample_ID != "P26208_1060_S62")

# Remove individual that was not vaccinated in dose 3 and dose 2
sampleTable <- subset(sampleTable, sample_ID != "P27910_1044_S44")
sampleTable <- subset(sampleTable, sample_ID != "P27910_1031_S31")
sampleTable <- subset(sampleTable, sample_ID != "P27910_1037_S37")

# test which one to remove to evaluate effect of single samples
#sampleTable <- sampleTable %>% filter(group != "pos" | batch != "third" | visit != 4 & visit != 5)
#test <- sampleTable %>% filter(group == "pos", batch == "third", visit == 5 | visit == 4) 
sampleTable <- sampleTable %>% filter(batch != "third")

#sampleTable <- sampleTable %>%
#  filter(!(sample_ID %in% (test %>% filter(!(test$subject_ID %in% unique(test$subject_ID)[c(2,4:6)])) %>% pull(sample_ID))))

#sampleTable <- sampleTable %>% filter(sample_ID %in% test$sample_ID)

# Remove corresponding samples from count table
#counts_raw <- counts_raw %>% select(-c(setdiff(colnames(counts_raw), sampleTable$sample_ID))) this one doesn't work
counts_raw <- subset(counts_raw, select = -c(P26208_1060_S62, P22761_1037_S37))


# Match count data with sample table
colnames(counts_raw) <- sampleTable$sample_ID
counts_raw <- counts_raw[, pmatch(sampleTable$sample_ID, colnames(counts_raw))]
all(rownames(sampleTable$sample_ID) == colnames(counts_raw))

# Read antibody data
antibody_data <- read.csv("../data/antibody_data/2022-04-26_antibody-binding-avidity.csv")

# Load BTM modules
gmt <- read.gmt("../data/blood_transcription_module/BTM_for_GSEA_20131008.gmt")

```


Removing reads with the log2 of the counts per million (cpm) lower than 1 (= 45896 genes). Plot genes detection rate across samples, comparing raw and filtered counts.

```{r}

meanLog2CPM <- rowMeans(log2(cpm(counts_raw) + 1))
counts_filtered <- counts_raw[meanLog2CPM > 1, ]

hist(rowSums(counts_raw > 3), main = title("Raw Counts"))
hist(rowSums(counts_filtered > 3), main = title("Filtered Counts"))

```


```

```

#DESeq 
```{r}

# Prepare for DESeq by creating new variables and factorizing 
sampleTable <- sampleTable %>%
  mutate(
    conditions = paste0(sampleTable$group, "_", sampleTable$dose),
    time = gsub(".{6}$", "", conditions),
    conditions = factor(conditions),
    subject_ID = factor(subject_ID),
    time = factor(time),
    sex = factor(sex),
    dose = factor(dose),
    group = factor(group),
    batch = factor(batch)
  )

# create 3 DESeq objects comparing against conditions, time and dose

dds1 <- DESeqDataSetFromMatrix(
  countData = round(counts_filtered),
  colData = as.data.frame(sampleTable),
  design = ~ sex + age + batch + conditions
)

dds2 <- DESeqDataSetFromMatrix(
  countData = round(counts_filtered),
  colData = as.data.frame(sampleTable),
  design = ~ sex + age + batch + time
)

dds3 <- DESeqDataSetFromMatrix(
  countData = round(counts_filtered),
  colData = as.data.frame(sampleTable),
  design = ~ sex + age + batch + dose
)


```
```
```
#DESeq experiments

```{r}
dds1 <-  estimateSizeFactors(dds1)

counts_vst <- assay(vst(dds1, blind = FALSE))

counts_filtered_vst <- counts_vst[, pmatch(sampleTable$sample_ID, colnames(counts_vst))]
#colnames(rpkm_filtered) <- sampleTable$sample_ID
counts_filtered_vst <- counts_filtered_vst[rowSums(counts_filtered_vst > 5) >= 32,]

all(rownames(sampleTable$sample_ID) == colnames(counts_filtered_vst))

colnames(counts_filtered_vst) <- plyr::mapvalues(colnames(counts_filtered_vst), from = sampleTable$sample_ID, to = paste(sampleTable$subject_ID, sampleTable$dose,sep="_"))

sampleTable <- sampleTable %>% mutate(ID_time_dose = paste(subject_ID, dose, sep = "_"))
rownames(sampleTable) <- sampleTable$ID_time_dose

# Run the DESeq2 analysis
dds1 <- DESeq(dds1, parallel = TRUE)
#resApeT <- lfcShrink(dds1, coef=2, type="apeglm", lfcThreshold=1)
#res <- results(resApeT)
#dds1 <- lfcShrink(dds1,  type = "apeglm", coef = 2,  lfcThreshold=1, res = results(dds1, contrast = c("conditions", "pos_24_dose2", "pos_0_dose2"), format = "DataFrame"))


dds2 <- DESeq(dds2, parallel = TRUE)
dds3 <- DESeq(dds3, parallel = TRUE)

# Contrast
resultsNames(dds1)
resultsNames(dds2)
resultsNames(dds3)

results(dds1)



counts_filtered_vst_alfa <- counts_filtered_vst[order(row.names(counts_filtered_vst)), ]





#chatgpts

# Set the seed for reproducibility
# set.seed(140)
# 
# # Extract the unique sample names (to_loop)
# to_loop <- unique(sub("_.*", "", colnames(counts_filtered_vst)))
# 
# # Function to calculate fold change
# calculate_fold_change <- function(column, counts_data) {
#   pattern_0_dose1 <- paste(column, "0_dose1", sep = "_")
#   pattern_24_dose1 <- paste(column, "24_dose1", sep = "_")
#   pattern_0_dose2 <- paste(column, "0_dose2", sep = "_")
#   pattern_24_dose2 <- paste(column, "24_dose2", sep = "_")
#   pattern_0_dose3 <- paste(column, "0_dose3", sep = "_")
#   pattern_24_dose3 <- paste(column, "24_dose3", sep = "_")
#   pattern_48_dose3 <- paste(column, "48_dose3", sep = "_")
# 
#   if (column == "222-21") {
#     fold_changes <- data.frame(
#       genes = rownames(counts_data),
#       FC_dose1 = counts_data[, pattern_24_dose1] / counts_data[, pattern_0_dose1],
#       FC_dose2 = NA
#     )
#   } else if (!pattern_0_dose3 %in% colnames(counts_data)) {
#     fold_changes <- data.frame(
#       genes = rownames(counts_data),
#       FC_dose1 = counts_data[, pattern_24_dose1] / counts_data[, pattern_0_dose1],
#       FC_dose2 = counts_data[, pattern_24_dose2] / counts_data[, pattern_0_dose2]
#     )
#   } else {
#     fold_changes <- data.frame(
#       genes = rownames(counts_data),
#       FC_dose1 = counts_data[, pattern_24_dose1] / counts_data[, pattern_0_dose1],
#       FC_dose2 = counts_data[, pattern_24_dose2] / counts_data[, pattern_0_dose2],
#       FC_dose3 = counts_data[, pattern_24_dose3] / counts_data[, pattern_0_dose3],
#       FC_dose3_48 = counts_data[, pattern_48_dose3] / counts_data[, pattern_0_dose3]
#     )
#   }
#   
#   return(fold_changes)
# }
# 
# # Calculate fold changes using lapply
# fold_changes_list <- lapply(to_loop, calculate_fold_change, counts_data = counts_filtered_vst)
# 
# # Combine the results into a data frame
# fold_changes <- data.table::rbindlist(fold_changes_list, idcol = TRUE, fill = TRUE) %>%
#   dplyr::rename(sample_ID = .id) %>%
#   tidyr::pivot_wider(names_from = sample_ID,
#                      values_from = c(FC_dose1, FC_dose2, FC_dose3, FC_dose3_48)) %>%
#   as.data.frame()
# 
# # Check the first few rows of the fold_changes data frame
# head(fold_changes)

set.seed(140)
to_loop <- unique(sub( "_.*","",colnames(counts_filtered_vst)))

df <- list()
for (column in to_loop){
  print(column)
  pattern_0_dose1 <- paste(column, "0_dose1",sep="_")
  pattern_24_dose1 <- paste(column, "24_dose1",sep="_")
  pattern_0_dose2 <- paste(column, "0_dose2",sep="_")
  pattern_24_dose2 <- paste(column, "24_dose2",sep="_")
  pattern_0_dose3 <- paste(column, "0_dose3",sep="_")
  pattern_24_dose3 <- paste(column, "24_dose3",sep="_")
  pattern_48_dose3 <- paste(column, "48_dose3",sep="_")
  
  if(column == "222-21"){
     df[[column]] <- data.frame(genes = rownames(counts_filtered_vst),
                                FC_dose1 = counts_filtered_vst[,pattern_24_dose1]/counts_filtered_vst[,pattern_0_dose1],
                                FC_dose2 = NA)
  }else if(!pattern_0_dose3 %in% colnames(counts_filtered_vst)){
    df[[column]] <- data.frame(genes = rownames(counts_filtered_vst),
                                FC_dose1 = counts_filtered_vst[,pattern_24_dose1]/counts_filtered_vst[,pattern_0_dose1],
                                FC_dose2 = counts_filtered_vst[,pattern_24_dose2]/counts_filtered_vst[,pattern_0_dose2])
  }else{
     df[[column]] <- data.frame(genes = rownames(counts_filtered_vst),
                                FC_dose1 = counts_filtered_vst[,pattern_24_dose1]/counts_filtered_vst[,pattern_0_dose1],
                                FC_dose2 = counts_filtered_vst[,pattern_24_dose2]/counts_filtered_vst[,pattern_0_dose2],
                                FC_dose3 =  counts_filtered_vst[,pattern_24_dose3]/counts_filtered_vst[,pattern_0_dose3]#,
                               # FC_dose3_48 = counts_filtered_vst[,pattern_48_dose3]/counts_filtered_vst[,pattern_0_dose3]
                               )
  }
}
  
fold_changes <- data.table::rbindlist(df, idcol = TRUE, fill = TRUE) %>%
  dplyr::rename(sample_ID = .id) %>%
  tidyr::pivot_wider(names_from = sample_ID,
                     values_from = c(FC_dose1,FC_dose2, FC_dose3#, FC_dose3_48
                                     )) %>%
  as.data.frame()

fold_changes <- data.frame(fold_changes, row.names = 1) %>%
  janitor::remove_empty(which = "cols") %>%
  as.matrix()

fold_changes <- fold_changes[order(row.names(fold_changes)), ]


# Log normalization
fold_changes <- log2(fold_changes)
# Z-score normalization and centering
fold_changes_norm <- t(scale(t(fold_changes)))
fold_changes_norm <- fold_changes_norm[!is.infinite(rowSums(fold_changes_norm)),]

metadata_2 <- sampleTable %>%
  mutate(heatmap_ID = paste0("FC", gsub("0|24","",dose),"_", gsub("-",".", subject_ID)),
         dose = gsub("0|24|_","",dose)) %>%
  dplyr::select(heatmap_ID,group, dose) %>%
  distinct()
rownames(metadata_2) <- metadata_2$heatmap_ID

# Generate new metadata containing subject ID and group only
metadata_3 <- metadata_2 %>%
  mutate(subject_ID = gsub(".*_","",heatmap_ID)) %>%
  tibble::rownames_to_column("samples") %>%
  dplyr::select(subject_ID, group) %>%
  distinct()

variable_to_predict <- "Spike_W6"
if(grepl("Avidity",variable_to_predict)){
  antibody_data_to_merge <- antibody_data %>%
  dplyr::select(sample_ID, all_of(variable_to_predict), prior_COVID.19)
  print("Avidity data detected, raw perentage valued used")
}else{
  antibody_data_to_merge <- antibody_data
  antibody_data_to_merge[[variable_to_predict]] <- log10(antibody_data[[variable_to_predict]])
  antibody_data_to_merge <- antibody_data_to_merge %>%
  dplyr::select(sample_ID, variable_to_predict, prior_COVID.19)
  print("Not avidity data, assuming it is antibody titer, thus the data is log10 normalized")
}
```

## Creating data tables...
```{r}
#btm_genes_names <- btm_modules %>% pull(`Module member genes`) %>% stringr::str_split(",") %>% unlist()
#btm_genes_names <- paste(btm_genes_names,collapse = "|")
top_1000 <- rownames(fold_changes)

fold_changes_ab_data <- fold_changes %>% 
  t() %>%
  as.data.frame() %>%
  # Select only the top variables that contribute to the first dimensions
  #select_if(colnames(.) %in% top_1000$name) %>%
  #select_if(colnames(.) %in% top_1000) %>%
  # Select only the genes from MO and IFN gene set modules
  #select_if(grepl(btm_genes_names,colnames(.))) %>%
  tibble::rownames_to_column("sample_ID") %>%
  tidyr::separate(sample_ID, into = c("X","dose", "subject_ID"), sep = "_") %>%
  tidyr::pivot_wider(names_from = "dose", values_from = c(everything(), -"subject_ID", -"X", -"dose")) %>%
  left_join(metadata_3,by = "subject_ID") %>%
  mutate(sample_ID = gsub("\\.","-",subject_ID)) %>%
  left_join(antibody_data_to_merge, by="sample_ID") %>%
  dplyr::select(-X) %>%
  mutate(avidity_classes = cut(.[[variable_to_predict]], 2, labels=FALSE),
         avidity_classes = paste0("class", avidity_classes)) %>%
  relocate(c("sample_ID","subject_ID","group",variable_to_predict,"avidity_classes" ))

Hfold_changes_ab_data <- fold_changes %>% as.data.frame() #%>%
 
  Hfold_changes_ab_data <- t(Hfold_changes_ab_data)
  Hfold_changes_ab_data <- as.data.frame(Hfold_changes_ab_data)
   Hfold_changes_ab_data <-   tibble::rownames_to_column(Hfold_changes_ab_data,"sample_ID")
  
   Hfold_changes_ab_data <- tidyr::separate(Hfold_changes_ab_data, sample_ID, into = c("X","dose", "subject_ID"), sep = "_")
  
   Hfold_changes_ab_data <- Hfold_changes_ab_data %>% slice(1:41)
   
   Hfold_changes_ab_data <- tidyr::pivot_wider(Hfold_changes_ab_data, names_from = 'dose', values_from = c(everything(), -"subject_ID", -"X", -"dose"))
  
  
# Select only the top variables that contribute to the first dimensions
  #select_if(colnames(.) %in% top_1000$name) %>%
  #select_if(colnames(.) %in% top_1000) %>%
  # Select only the genes from MO and IFN gene set modules
  #select_if(grepl(btm_genes_names,colnames(.))) %>%
  tibble::rownames_to_column("sample_ID") %>%
  tidyr::separate(sample_ID, into = c("X","dose", "subject_ID"), sep = "_") %>%
  tidyr::pivot_wider(names_from = "dose", values_from = c(everything(), -"subject_ID", -"X", -"dose")) %>%
  left_join(metadata_3,by = "subject_ID") %>%
  mutate(sample_ID = gsub("\\.","-",subject_ID)) %>%
  left_join(antibody_data_to_merge, by="sample_ID") %>%
  dplyr::select(-X) %>%
  mutate(avidity_classes = cut(.[[variable_to_predict]], 2, labels=FALSE),
         avidity_classes = paste0("class", avidity_classes)) %>%
  relocate(c("sample_ID","subject_ID","group",variable_to_predict,"avidity_classes" ))










unnest(fold_changes_ab_data, A1BG_dose1) 


#as.data.table(fold_changes_ab_data)[, unlist(fold_changes_ab_data[,6], by = list(row, col1)]$col1




fold_changes_ab_spearman_data <- fold_changes_norm %>% 
  t() %>%
  as.data.frame() %>%
  # Select only the top variables that contribute to the first dimensions
  #select_if(colnames(.) %in% top_1000$name) %>%
  #select_if(colnames(.) %in% top_1000$name) %>%
  # Select only the genes from MO and IFN gene set modules
  #select_if(grepl(btm_genes_names,colnames(.))) %>%
  tibble::rownames_to_column("sample_ID") %>%
  tidyr::separate(sample_ID, into = c("X","dose", "subject_ID"), sep = "_") %>%
  tidyr::pivot_wider(names_from = "dose", values_from = c(everything(), -"subject_ID", -"X", -"dose")) %>%
  left_join(metadata_3,by = "subject_ID") %>%
  mutate(sample_ID = gsub("\\.","-",subject_ID)) %>%
  left_join(antibody_data_to_merge, by="sample_ID") %>%
  dplyr::select(-X) %>%
  mutate(avidity_classes = cut(.[[variable_to_predict]], 2, labels=FALSE),
         avidity_classes = paste0("class", avidity_classes)) %>%
  relocate(c("sample_ID","subject_ID","group",variable_to_predict,"avidity_classes" ))






# fold_changes_ab_data <- fold_changes_ab_data %>% slice(1:15)
# fold_changes_ab_spearman_data <- fold_changes_ab_spearman_data %>% slice(1:15)
# 
# 
# 
# 
# 
# 
# fold_changes_ab_data <- apply(fold_changes_ab_data,6,as.character)
# fold_changes_ab_data <- as.data.frame(fold_changes_ab_data)
# 
# 
# fold_changes_ab_datax = fold_changes_ab_data %>% as.data.frame(6:length(fold_changes_ab_data))
# 
# fold_changes_ab_datax <- as.data.frame(fold_changes_ab_datax)
# 
# 
# foo = data.frame(col1 = c(1,2,3)) %>% mutate(col2 = list("a", "b", "c") %>% unlist())
# 
# 
# deg_ds_d3 <- dplyr::select(fold_changes_ab_data, contains('AIM2'))


 data.table::fwrite(fold_changes_ab_data, "../data/machine_learning_dataset/fold_changes_mofa.csv", row.names = F, sep = ",", na = NA, nThread = 8)

 # 
 data.table::fwrite(fold_changes_ab_spearman_data, "../data/machine_learning_dataset/fold_changes_correlation.csv", row.names = F, sep = ",",nThread = 8)

write.csv(fold_changes_ab_data, file = "../data/machine_learning_dataset/fold_changes_mofa.csv", na = "NA")


# 
# # Assuming you have a DESeqDataSet object called 'dds'
# 
# # 1. Subset the DESeq data set
# # Define the relevant columns for time points (e.g., 0h and 24h)
# time_points <- c("0h", "24h")
# 
# # Define the relevant doses (e.g., dose 1, dose 2, dose 3)
# doses <- c("dose1", "dose2", "dose3")
# 
# # Create an empty data frame to store the results
# fold_changes <- data.frame()
# 
# # Loop through each dose
# for (dose in doses) {
#   # Subset the DESeq data for the current dose
#   dose_data <- assay(dds)[, colData(dds)$dose == dose]
# 
#   # Loop through each gene
#   for (gene_id in rownames(dose_data)) {
#     # Calculate the fold change for each gene
#     fc_values <- dose_data[gene_id, time_points[2]] / dose_data[gene_id, time_points[1]]
# 
#     # Handle missing data by substituting NA values
#     if (any(is.na(fc_values))) {
#       fc_values[is.na(fc_values)] <- NA
#     }
# 
#     # Create a data frame for the fold change values
#     fc_df <- data.frame(
#       Gene = gene_id,
#       Dose = dose,
#       FoldChange = fc_values
#     )
# 
#     # Append the results to the fold_changes data frame
#     fold_changes <- rbind(fold_changes, fc_df)
#   }
# }
# 
# # 4. Organize the results in a structured data frame
# # View the calculated fold changes
# print(fold_changes)
```

## Calculating Fold Change by TPM: Loading data
```{r}
# Load raw counts

# Filter for only protein coding genes or mitochondrial/ribosomic genes
# gene_id_metadata <- read.csv("../data/metadata/2023-11-222_mart_export_gene_ids_hsa.txt") %>% filter(Gene.type %in% c("Mt_rRNA", "Mt_rRNA", "protein_coding", "rRNA"))

gene_id_metadata <- read.csv("../data/metadata/2023-11-222_mart_export_gene_ids_hsa.txt") %>% filter(Gene.type %in% c("protein_coding"))

tpm_gene <- data.table::fread("../data/processed_data_nfcore/batch_1-2/salmon/salmon.merged.gene_tpm.tsv",  header = TRUE, sep = "\t") %>% 
  as.data.frame()

tpm_fc <- tpm_gene

#data_tpm1 <- tpm_fc

colnames(tpm_gene) <- sub("_L001|_L002", "",colnames(tpm_gene))

# Identify duplicated gene_names
duplicated_genes <- tpm_gene$gene_name[duplicated(tpm_gene$gene_name)]

# Replace duplicated gene_names with corresponding gene_ids /
tpm_gene$gene_name <- ifelse(tpm_gene$gene_name %in% duplicated_genes,
                               tpm_gene$gene_id,
                               tpm_gene$gene_name)

# Change row names to gene names
rownames(tpm_gene) <- tpm_gene$gene_name
tpm_gene <- dplyr::select(tpm_gene, -c(gene_name, gene_id))
```

## New chunk for updated filtering code

```{r}

# Load raw counts
tpm_raw <- data.table::fread("../data/processed_data_nfcore/batch_1-2/salmon/salmon.merged.gene_tpm.tsv",
                                  header = TRUE, sep = "\t") %>% 
  as.data.frame() %>%
  select(-gene_id)


colnames(tpm_raw) <- sub("_L001|_L002", "",colnames(tpm_raw))


# Aggregate and sum up genes with same gene symbol (sum up isoform of genes)
tpm_raw <- tpm_raw[tpm_raw$gene_name %in% gene_id_metadata$Gene.name,]

tpm_raw <- aggregate(tpm_raw[, -1], 
                        list(gene_name = tpm_raw[, 1]), 
                        FUN = sum)


# Change row names to gene names
rownames(tpm_raw) <- tpm_raw$gene_name
tpm_raw <- dplyr::select(tpm_raw, -c(gene_name))

tpm_gene <- tpm_raw
```

## Data analysis 
```{r}
# Remove corresponding samples from count table /Originally in Metadata chunk
tpm_gene <- subset(tpm_gene, select = -c(P26208_1060_S62, P22761_1037_S37))

# Match count data with sample table
colnames(tpm_gene) <- sampleTable$sample_ID
tpm_gene <- tpm_gene[, pmatch(sampleTable$sample_ID, colnames(tpm_gene))]
all(rownames(sampleTable$sample_ID) == colnames(tpm_gene))




 #colnames(data_tpm1) <- plyr::mapvalues(sub("_L001|_L002","",colnames(tpm_gene)), from = sampleTable$sample_ID, to = paste(sampleTable$subject_ID,sampleTable$dose, sampleTable$group, sampleTable$batch, sep = "_")) 

cxcl10_data <- dplyr::filter(tpm_gene, grepl('CXCL10',rownames(tpm_gene)))


# Use table() to count occurrences of each gene name
#gene_counts <- table(tpm_fc$gene_name)

# Convert the result to a data frame if needed
#gene_counts_df <- as.data.frame(gene_counts)

# Rename columns for clarity (optional)
#colnames(gene_counts_df) <- c("GeneName", "Count")

# View the resulting data frame with gene counts
#print(gene_counts_df)
```

## Filtering low counts

Removing reads with the log2 of the counts per million (cpm) lower than 1 (= 45896 genes). Plot genes detection rate across samples, comparing raw and filtered counts.
```{r}
#meanLog2TPM <- rowMeans(log2(cpm(tpm_gene) + 1))
meanLog2TPM <- rowMeans(log2(tpm_gene + 1))
tpm_filtered <- tpm_gene[meanLog2TPM > 0.1, ]


#`ASAP2`, `ATP7B`, `BRCA2`, `C3`,  `C5`, `CACNA1A`,`FMNL2`,  `IL31RA`,  `MGAM2`,   `SIGLEC11`,   `UBQLNL`

genes10_data <- dplyr::filter(tpm_filtered, grepl('UBQLNL',rownames(tpm_filtered))) 
cxcl10_data <- dplyr::filter(tpm_filtered, grepl('CXCL10',rownames(tpm_filtered)))

hist(rowSums(tpm_gene > 3), main = title("TPM Counts"))
hist(rowSums(tpm_filtered > 3), main = title("Filtered TPM Counts"))

```

##Calculating TPM
```{r}

counts_filtered_tpm <- tpm_filtered[, pmatch(sampleTable$sample_ID, colnames(tpm_filtered))]
#colnames(rpkm_filtered) <- sampleTable$sample_ID
counts_filtered_tpm <- counts_filtered_tpm[rowSums(counts_filtered_tpm > 0) >= 0,]

#cxcl_data <- dplyr::select(counts_filtered_tpm, contains('CXCL10'))

cxcl_data <- dplyr::filter(counts_filtered_tpm, grepl('CXCL10',rownames(counts_filtered_tpm))) 


all(rownames(sampleTable$sample_ID) == colnames(counts_filtered_tpm))

colnames(counts_filtered_tpm) <- plyr::mapvalues(colnames(counts_filtered_tpm), from = sampleTable$sample_ID, to = paste(sampleTable$subject_ID, sampleTable$dose,sep="_"))

sampleTable <- sampleTable %>% mutate(ID_time_dose = paste(subject_ID, dose, sep = "_"))
rownames(sampleTable) <- sampleTable$ID_time_dose

#pseudocount = 0.8
counts_filtered_tpm <- counts_filtered_tpm %>% mutate(counts_filtered_tpm + 0.8)

#Checking FC calculations prior to for-loop
aatk_14_d1 <- counts_filtered_tpm[15,5]/counts_filtered_tpm[15,1]
aatk_8_d2 <- counts_filtered_tpm[15,16]/counts_filtered_tpm[15,12]

zyx_14_d1 <- counts_filtered_tpm[11894,5]/counts_filtered_tpm[11894,1]
zyx_8_d2 <- counts_filtered_tpm[11894,16]/counts_filtered_tpm[11894,12]


set.seed(140)
to_loop <- unique(sub( "_.*","",colnames(counts_filtered_tpm)))

df <- list()
for (column in to_loop){
  print(column)
  pattern_0_dose1 <- paste(column, "0_dose1",sep="_")
  pattern_24_dose1 <- paste(column, "24_dose1",sep="_")
  pattern_0_dose2 <- paste(column, "0_dose2",sep="_")
  pattern_24_dose2 <- paste(column, "24_dose2",sep="_")
  pattern_0_dose3 <- paste(column, "0_dose3",sep="_")
  pattern_24_dose3 <- paste(column, "24_dose3",sep="_")
  pattern_48_dose3 <- paste(column, "48_dose3",sep="_")
  
  if(column == "222-21"){
     df[[column]] <- data.frame(genes = rownames(counts_filtered_tpm),
                                FC_dose1 = counts_filtered_tpm[,pattern_24_dose1]/counts_filtered_tpm[,pattern_0_dose1],
                                FC_dose2 = NA)
  }else if(!pattern_0_dose3 %in% colnames(counts_filtered_tpm)){
    df[[column]] <- data.frame(genes = rownames(counts_filtered_tpm),
                                FC_dose1 = counts_filtered_tpm[,pattern_24_dose1]/counts_filtered_tpm[,pattern_0_dose1],
                                FC_dose2 = counts_filtered_tpm[,pattern_24_dose2]/counts_filtered_tpm[,pattern_0_dose2])
  }else{
     df[[column]] <- data.frame(genes = rownames(counts_filtered_tpm),
                                FC_dose1 = counts_filtered_tpm[,pattern_24_dose1]/counts_filtered_tpm[,pattern_0_dose1],
                                FC_dose2 = counts_filtered_tpm[,pattern_24_dose2]/counts_filtered_tpm[,pattern_0_dose2],
                                FC_dose3 =  counts_filtered_tpm[,pattern_24_dose3]/counts_filtered_tpm[,pattern_0_dose3]#,
                               # FC_dose3_48 = counts_filtered_vst[,pattern_48_dose3]/counts_filtered_vst[,pattern_0_dose3]
                               )
  }
}
  
fold_changes_tpm <- data.table::rbindlist(df, idcol = TRUE, fill = TRUE) %>%
  dplyr::rename(sample_ID = .id) %>%
  tidyr::pivot_wider(names_from = sample_ID,
                     values_from = c(FC_dose1,FC_dose2, FC_dose3#, FC_dose3_48
                                     )) %>%
  as.data.frame()

fold_changes_tpm <- data.frame(fold_changes_tpm, row.names = 1) %>%
  janitor::remove_empty(which = "cols") %>%
  as.matrix()

fold_changes_tpm <- fold_changes_tpm[order(row.names(fold_changes_tpm)), ]




# Log normalization
fold_changes_tpm <- log2(fold_changes_tpm)
# Z-score normalization and centering
fold_changes_norm_tpm <- t(scale(t(fold_changes_tpm)))
fold_changes_norm_tpm <- fold_changes_norm_tpm[!is.infinite(rowSums(fold_changes_norm_tpm)),]

metadata_2 <- sampleTable %>%
  mutate(heatmap_ID = paste0("FC", gsub("0|24","",dose),"_", gsub("-",".", subject_ID)),
         dose = gsub("0|24|_","",dose)) %>%
  dplyr::select(heatmap_ID,group, dose) %>%
  distinct()
rownames(metadata_2) <- metadata_2$heatmap_ID

# Generate new metadata containing subject ID and group only
metadata_3 <- metadata_2 %>%
  mutate(subject_ID = gsub(".*_","",heatmap_ID)) %>%
  tibble::rownames_to_column("samples") %>%
  dplyr::select(subject_ID, group) %>%
  distinct()

variable_to_predict <- "Spike_W6"
if(grepl("Avidity",variable_to_predict)){
  antibody_data_to_merge <- antibody_data %>%
  dplyr::select(sample_ID, all_of(variable_to_predict), prior_COVID.19)
  print("Avidity data detected, raw perentage valued used")
}else{
  antibody_data_to_merge <- antibody_data
  antibody_data_to_merge[[variable_to_predict]] <- log10(antibody_data[[variable_to_predict]])
  antibody_data_to_merge <- antibody_data_to_merge %>%
  dplyr::select(sample_ID, variable_to_predict, prior_COVID.19)
  print("Not avidity data, assuming it is antibody titer, thus the data is log10 normalized")
}

fold_changes_ab_TPM_data <- fold_changes_tpm %>% 
  t() %>%
  as.data.frame() %>%
  # Select only the top variables that contribute to the first dimensions
  #select_if(colnames(.) %in% top_1000$name) %>%
  #select_if(colnames(.) %in% top_1000) %>%
  # Select only the genes from MO and IFN gene set modules
  #select_if(grepl(btm_genes_names,colnames(.))) %>%
  tibble::rownames_to_column("sample_ID") %>%
  tidyr::separate(sample_ID, into = c("X","dose", "subject_ID"), sep = "_") %>%
  tidyr::pivot_wider(names_from = "dose", values_from = c(everything(), -"subject_ID", -"X", -"dose")) %>%
  left_join(metadata_3,by = "subject_ID") %>%
  mutate(sample_ID = gsub("\\.","-",subject_ID)) %>%
  left_join(antibody_data_to_merge, by="sample_ID") %>%
  dplyr::select(-X) %>%
  mutate(avidity_classes = cut(.[[variable_to_predict]], 2, labels=FALSE),
         avidity_classes = paste0("class", avidity_classes)) %>%
  relocate(c("sample_ID","subject_ID","group",variable_to_predict,"avidity_classes" ))

fold_changes_ab_spearman_TPM_data <- fold_changes_norm_tpm %>% 
  t() %>%
  as.data.frame() %>%
  # Select only the top variables that contribute to the first dimensions
  #select_if(colnames(.) %in% top_1000$name) %>%
  #select_if(colnames(.) %in% top_1000$name) %>%
  # Select only the genes from MO and IFN gene set modules
  #select_if(grepl(btm_genes_names,colnames(.))) %>%
  tibble::rownames_to_column("sample_ID") %>%
  tidyr::separate(sample_ID, into = c("X","dose", "subject_ID"), sep = "_") %>%
  tidyr::pivot_wider(names_from = "dose", values_from = c(everything(), -"subject_ID", -"X", -"dose")) %>%
  left_join(metadata_3,by = "subject_ID") %>%
  mutate(sample_ID = gsub("\\.","-",subject_ID)) %>%
  left_join(antibody_data_to_merge, by="sample_ID") %>%
  dplyr::select(-X) %>%
  mutate(avidity_classes = cut(.[[variable_to_predict]], 2, labels=FALSE),
         avidity_classes = paste0("class", avidity_classes)) %>%
  relocate(c("sample_ID","subject_ID","group",variable_to_predict,"avidity_classes" ))


 data.table::fwrite(fold_changes_ab_TPM_data, "../data/machine_learning_dataset/TPM_fold_changes_omics_protein_coding_3.csv", row.names = F, sep = ",", na = NA, nThread = 8)

 # 
 data.table::fwrite(fold_changes_ab_spearman_TPM_data, "../data/machine_learning_dataset/TPM_fold_changes_correlation_protein_coding_3.csv", row.names = F, sep = ",",nThread = 8)


```

