---
title: "MixOmics Örebro Study"
author: "Rodrigo Arcoverde & Matthew Hinchcliff"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output: html_document
abstract: Short project description.
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/lab_book/")})
---

```{r setup, include=FALSE}
knitr::opts_knit$set(echo = TRUE,
                     root.dir = getwd(),
                     fig.width = 6, fig.height = 5,
                     warning = FALSE, 
                     message = FALSE)

result.dir <- paste0("results/",Sys.Date(),"/")
diablo_figures.dir <- paste0("results/", Sys.Date(), "/", "diablo_figures/")

## creates result.dir with date in if not existent
ifelse(isFALSE(dir.exists(paste0("../",result.dir))), dir.create(paste0("../",result.dir),recursive = TRUE),"Result directory for today exists already!")

ifelse(isFALSE(dir.exists(paste0("../", diablo_figures.dir))),
  dir.create(paste0("../", diablo_figures.dir), recursive = TRUE),
  "Result directory for today exists already!"
)

options(stringsAsFactors = FALSE) 


```

## Needed libraries
```{r libraries, message=FALSE}

library(mixOmics)
library(data.table)
library(kableExtra)
library(forcats)
library(tibble)
library(devtools)
library(tidyr)
library(dplyr)
library(plyr)
library(SummarizedExperiment)
library(grDevices)
library(ggplot2)
library(ggpubr)
library(ggprism)
```

## Load data
```{r message=FALSE, warning=FALSE}

# Read the transcript data
transcript_data <- data.table::fread("../data/machine_learning_dataset/fold_changes_correlation.csv") 

transcript_data_tpm <- data.table::fread("../data/machine_learning_dataset/TPM_fold_changes_correlation_protein_coding.csv") 

# Read the cytokine data
cytokine_data <- data.table::fread("../data/cytokine_luminex/220406_luminex_edited_data_FC.csv")

set.seed(5249) # for reproducibility, remove for normal use

#For using TPM, removes Prior covid_19 column
transcript_data <- transcript_data_tpm[,1:39599]

```


## Take environment snapshot 
```{r}
#renv::snapshot()
```

## Metadata
```{r}

# Load sample metadata
sampleTable <- read.table("../data/metadata/sample_metadata_3.csv", 
                          sep = ",", header = TRUE) %>%
  dplyr::select(-c(age, sex))

# Load patient metadata
patient_metadata <- read.table("../data/metadata/patient_metadata.csv", 
                               sep = ",", header = TRUE)

# Join sample and patient metadata
sampleTable <- sampleTable %>% left_join(patient_metadata, by = "subject_ID")

# Change visits to dose and time point...
sampleTable <- sampleTable %>%
  mutate(
    dose = as.factor(plyr::mapvalues(visit,
      from = c(1, 2, 4, 5, 10, 11, 12),
      to   = c("0_dose1", "24_dose1", "0_dose2", "24_dose2", "0_dose3", "24_dose3", "48_dose3")
    )),
    group = as.factor(group)
  ) %>%
  # ...(sex) 0 and 1 to male and female...
  mutate(
    sex = as.factor(plyr::mapvalues(sex,
      from = c(0, 1),
      to   = c("male", "female")
    )),
    group = as.factor(sex)
  ) %>%
  # ...(prior covid) 0 and 1 to neg and pos
  mutate(
    prior_covid19 = as.factor(plyr::mapvalues(prior_covid19,
      from = c(0, 1),
      to   = c("neg", "pos")
    )),
    group = as.factor(prior_covid19)
  )

# Remove 24dose_2 from subject 21 (missing 0dose_2)
sampleTable <- subset(sampleTable, sample_ID != "P22761_1037_S37")

# Remove one sample from subject 7 (sample replicate)
sampleTable <- subset(sampleTable, sample_ID != "P26208_1060_S62")

# Remove individual that was not vaccinated in dose 3 and dose 2
sampleTable <- subset(sampleTable, sample_ID != "P27910_1044_S44")
sampleTable <- subset(sampleTable, sample_ID != "P27910_1031_S31")
sampleTable <- subset(sampleTable, sample_ID != "P27910_1037_S37")

# test which one to remove to evaluate effect of single samples
#sampleTable <- sampleTable %>% filter(group != "pos" | batch != "third" | visit != 4 & visit != 5)
#test <- sampleTable %>% filter(group == "pos", batch == "third", visit == 5 | visit == 4) 
sampleTable <- sampleTable %>% filter(batch != "third")

# Aligns transcript data & cytokine data by sample_ID

transcript_data <- arrange(transcript_data, sample_ID)
cytokine_data <- arrange(cytokine_data, subject_ID)

transcript_data_tpm <- arrange(transcript_data_tpm, sample_ID)
```

## Data Pre-processing
```{r}
#function definitions
not_all_ones <- function(x) {
  x %>%
    select(which(colSums(. == 1, na.rm = TRUE) < nrow(.)))
}

#NA removal function
not_all_na <- function(x) any(!is.na(x))

#Removing irrelevant columns
trans_ds <- select(transcript_data, c(1, 3, 6:length(transcript_data))) #selects the sample_ID, group, and gene data 
trans_ds_tpm <- select(transcript_data_tpm, c(1, 3, 6:length(transcript_data_tpm))) #selects the sample_ID, group, and gene data 
trans_ds_tpm_na <- na.omit(trans_ds_tpm)

cyto_ds <- select(cytokine_data, c(1, 3:35)) # Selects sample_ID, group, and cytokine data


#Resetting group values & names
cyto_ds[cyto_ds == "Experienced"] <- "pos"
cyto_ds[cyto_ds == "Naive"] <- "neg"
cyto_ds <- plyr::rename(cyto_ds, c("subject_ID" = "sample"))
cyto_ds <- plyr::rename(cyto_ds, c("Study group" = "group"))  
  
#Matches samples against against the transcript data
names_match <- trans_ds$sample_ID
cyto_ds <- cyto_ds %>%
  dplyr::filter(sample %in% names_match)

#Creates a list of exp/naive matching the sample
trans_groups <- as.data.frame(trans_ds$sample_ID)
trans_groups <- cbind(trans_groups,trans_ds$group)
trans_groups <- plyr::rename(trans_groups, c("trans_ds$group" = "group", "trans_ds$sample_ID" = "sample"))
trans_groups_exp <- trans_groups %>% filter(group == "pos")
trans_groups_nai <- trans_groups %>% filter(group == "neg")

#Doses 1-3
tds <- column_to_rownames(trans_ds, var = "sample_ID")
#Filtering into separate objects based on dose
transcript_d1 <- select(tds, contains('_dose1'))
transcript_d2 <- select(tds, contains('_dose2'))
transcript_d3 <- select(tds, contains('_dose3'))

#Removing sample 222-15,  222-2, and 222-21 for symmetrical matrices for analyses
transcript_stack1 <- transcript_d1[-c(4,7:8),]
transcript_stack1$dose <- paste(dose = '1')
transcript_stack1 <- rownames_to_column(transcript_stack1, var = 'sample')
transcript_stack1$sample <-  paste(transcript_stack1$sample, transcript_stack1$dose, sep = ".")

transcript_stack2 <- transcript_d2[-c(4,7:8),]
transcript_stack2$dose <- paste(dose = '2')
transcript_stack2 <- rownames_to_column(transcript_stack2, var = 'sample')
transcript_stack2$sample <-  paste(transcript_stack2$sample, transcript_stack2$dose, sep = ".")

transcript_stack3 <- transcript_d3[-c(4,7:8),]
transcript_stack3$dose <- paste(dose = '3')
transcript_stack3 <- rownames_to_column(transcript_stack3, var = 'sample')
transcript_stack3$sample <-  paste(transcript_stack3$sample, transcript_stack3$dose, sep = ".")

# Define the string you want to remove
string_to_remove <- "_dose1"
string_to_remove2 <- "_dose2"
string_to_remove3 <- "_dose3"

# Remove the string from column names
colnames(transcript_stack1) <- sub(string_to_remove, "", colnames(transcript_stack1))
colnames(transcript_stack2) <- sub(string_to_remove2, "", colnames(transcript_stack2))
colnames(transcript_stack3) <- sub(string_to_remove3, "", colnames(transcript_stack3))

transcript_stack_all <- transcript_stack1 %>% bind_rows(transcript_stack2, transcript_stack3)


#Dose 1 
transcript_d1g <- bind_cols(transcript_d1, trans_groups)

#Dose 2 removing patient 222-21
transcript_d2g <- bind_cols(transcript_d2, trans_groups)
#transcript_d2g <- rownames_to_column(transcript_d2g, var = "sample")
transcript_d2 <- na.omit(transcript_d2)
transcript_d2g <- na.omit(transcript_d2g)

#Dose 3 removing NAs
transcript_d3g <- bind_cols(transcript_d3, trans_groups)
#transcript_d3g <- rownames_to_column(transcript_d3g, var = "sample")
transcript_d3 <- na.omit(transcript_d3)
transcript_d3g <- na.omit(transcript_d3g)

transcript_stack <- transcript_d1[-c(4,7:8),]
transcript_stack$dose <- paste(dose = '1')
transcript_stack <- rownames_to_column(transcript_stack, var = 'sample')
transcript_stack$sample <-  paste(transcript_stack$sample, transcript_stack$dose, sep = ".")

transcript_stack2 <- transcript_d2[-c(4,7:8),]
transcript_stack2$dose <- paste(dose = '2')
transcript_stack2 <- rownames_to_column(transcript_stack2, var = 'sample')
transcript_stack2$sample <-  paste(transcript_stack2$sample, transcript_stack2$dose, sep = ".")

transcript_stack2 <- transcript_d2[-c(4,7:8),]
transcript_stack2$dose <- paste(dose = '2')
transcript_stack2 <- rownames_to_column(transcript_stack2, var = 'sample')
transcript_stack2$sample <-  paste(transcript_stack2$sample, transcript_stack2$dose, sep = ".")


#Dose 1 cytokine data
time_d1 <- "V2"
cyto_d1 <- cyto_ds %>% dplyr::filter(Timepoint %in% time_d1)
#Dose 1 Cytokine correlation object 
cyto_d1 <- select(cyto_d1, c(4:34)) 
cyto_d1 <- not_all_ones(cyto_d1)
cyto_d1 <- as.data.frame(cyto_d1) %>% select(where(not_all_na))


#Dose 2 cytokine data
time_d2 <- "V5"
cyto_d2 <- cyto_ds %>% dplyr::filter(Timepoint %in% time_d2)
#Cytokine correlation data dose
cyto_d2 <- select(cyto_d2, c(4:34)) 
#removing 222-21
cyto_d2 <- cyto_d2[-c(8),]
cyto_d2 <- not_all_ones(cyto_d2)
cyto_d2 <- as.data.frame(cyto_d2) %>% select(where(not_all_na))



#Dose 3 cytokine data
time_d3 <- "V11"
cyto_d3 <- cyto_ds %>% dplyr::filter(Timepoint %in% time_d3)
#Dose 3 Cytokine correlation object 
cyto_d3 <- select(cyto_d3, c(4:34))
#removing samples with missing transcript data
cyto_d3 <- cyto_d3[-c(4,7:8),]
cyto_d3 <- as.data.frame(cyto_d3) %>% select(where(not_all_na))
cyto_d3 <- na.omit(cyto_d3)
cyto_d3 <- not_all_ones(cyto_d3)


#All Doses
#removes NAs
td <- tds[-c(4,7:8),2:length(tds)]
trans_groups_alldoses <- trans_groups[-c(4,7:8),] 

#Filters by dose
time <- c("V11","V5", "V2")
cyto <- cyto_ds %>% dplyr::filter(Timepoint %in% time)

# Removed NAs and columns where all values are 1 manually for PCA analysis with all doses, as well as dose-wide comparisons due
# to reduced cytokine panel after dose 3
cyto <- cyto[,-c(5,7,9:15,17,19,25,30,31,33,34)]
cyto_groups_time <- cyto[-c(12,21),c(1:3)]
cyto$sample <- paste(cyto$sample, cyto$Timepoint, sep = ".")
cyto_groups <- cyto[-c(12,21),c(1,3)]
cyto <- column_to_rownames(cyto, var = 'sample')
cyto_data <- select(cyto, c(4:length(cyto))) 

cyto_data <- cyto_data[-c(12,21),]
cyto_data <- as.data.frame(cyto_data) %>% select(where(not_all_na))
cyto_data <- na.omit(cyto_data)
cyto_data <- not_all_ones(cyto_data)

cyto_groups <- cyto[-c(12,21),c(1,3)]

```

## PCA
```{r}

#All doses by group

#removing zero variance genes: -[,13809] H2AC19_dose3
td <- td[,-13809]

pca.gene <- pca(td, ncomp = 10, center = TRUE, scale = TRUE)
pca.cyto <- pca(cyto_data, ncomp = 10, center =T, scale = T)
pca.gene_all <- pca(transcript_stack_all, ncomp = 10, center = TRUE, scale = TRUE)

plot(pca.gene)
plot(pca.cyto, )
plot(pca.gene_all)

plotIndiv(pca.gene, comp = c(1, 2), 
              group = trans_groups_alldoses$group, 
              ind.names = trans_groups_alldoses$sample, 
              legend = TRUE, title = 'All Doses mRNA transcripts by group, PCA comp 1 - 2')

## This one needs an object from a later data object. 
plotIndiv(pca.gene_all, comp = c(1, 2),
              group = dia_group$group,
              ind.names = F,
              legend = TRUE, title = 'All Doses mRNA transcripts by group, PCA comp 1 - 2')


plotIndiv(pca.gene_all, comp = c(1, 2),
          group = transcript_stack_all$dose,
          ind.names = transcript_stack_all$sample,
          legend = TRUE, title = 'All Doses mRNA transcripts by dose, PCA comp 1 - 2')



plotIndiv(pca.cyto, comp = c(1, 2), 
          group = cyto_groups_time$group, 
          ind.names = cyto_groups_time$sample, 
          legend = TRUE, title = 'Cytokines All doses by group, PCA comp 1 - 2')


{
  pdf(paste0("../",diablo_figures.dir,"PCA_mRNA_by_group.pdf"),  width = 10, height = 10)
  plotIndiv(pca.gene, comp = c(1, 2), 
              group = trans_groups_alldoses$group, 
              ind.names = trans_groups_alldoses$sample, 
              legend = TRUE, title = 'All Doses mRNA transcripts by group, PCA comp 1 - 2')
  dev.off()
}

{
  pdf(paste0("../",diablo_figures.dir,"PCA_mRNA_by_group_all_doses.pdf"),  width = 10, height = 10)
  plotIndiv(pca.gene_all, comp = c(1, 2),
              group = dia_group$group,
              ind.names = F,
              legend = TRUE, title = 'All Doses mRNA transcripts by group, PCA comp 1 - 2') 

  dev.off()
}



{
  pdf(paste0("../",diablo_figures.dir,"PCA_cytokines_by_group.pdf"),  width = 10, height = 10)
  plotIndiv(pca.cyto, comp = c(1, 2), 
          group = cyto_groups_time$group, 
          ind.names = F, 
          legend = TRUE, title = 'Cytokines All doses by group, PCA comp 1 - 2')
  dev.off()
}


# Extract the loadings for the first and second principal components
loadings_pc1_gene <- pca.gene$rotation[, 1]
#loadings_pc2_gene <- pca.gene$rotation[, 2]
#top_pca_gene <- pca.gene$
#print(top_pca_gene)

# Obtain the top ten loadings for PC1 and PC2
top_ten_pc1_gene <- head(sort(abs(loadings_pc1_gene), decreasing = TRUE), 10)
#top_ten_pc2_gene <- head(sort(abs(loadings_pc2_gene), decreasing = TRUE), 10)

# Display the top ten loadings for PC1 and PC2
top_ten_pc1_gene
#top_ten_pc2_gene


# Extract the loadings for the first and second principal components
loadings_pc1_cyto <- pca.cyto$rotation[, 1]
#loadings_pc2_cyto <- pca.cyto$rotation[, 2]

# Obtain the top ten loadings for PC1 and PC2
top_5_pc1_cyto <- head(sort(abs(loadings_pc1_cyto), decreasing = TRUE), 5)
#top_ten_pc2_cyto <- head(sort(abs(loadings_pc2_cyto), decreasing = TRUE), 10)

# Display the top ten loadings for PC1 and PC2
top_5_pc1_cyto
#top_ten_pc2_cyto

# Extract the loadings for the first and second principal components
loadings_pc1_gene_all <- pca.gene_all$rotation[, 1]
#loadings_pc2_gene_all <- pca.gene_all$rotation[, 2]

# Obtain the top ten loadings for PC1 and PC2
top_ten_pc1_gene_all <- head(sort(abs(loadings_pc1_gene_all), decreasing = TRUE), 10)
#top_ten_pc2_gene_all <- head(sort(abs(loadings_pc2_gene_all), decreasing = TRUE), 10)

# Display the top ten loadings for PC1 and PC2
top_ten_pc1_gene_all
#top_ten_pc2_gene_all


```




## Constructing DIABLO objects Cytokines and Genes

```{r}
set.seed(5249) # for reproducibility, remove for normal use


#Filtering cytokines to remove NAs and columns where all values are 1
cyto_data_stack <- cyto_data[-c(10:11,18:22),]
transcript_stack_all_spls <- column_to_rownames(transcript_stack_all, var = "sample")
#removing dose sorting column
transcript_stack_all_spls <- transcript_stack_all_spls[,-13198]
#transcript_stack_all_spls <- transcript_stack_all_spls[,-6169]




```

## mDIABLO MixOmics integration, data pre-processing 
```{r}

set.seed(123) # for reproducibility, remove for normal use
#Data pre-processing for mDIABLO data visualization 
all_doses <- rownames_to_column(transcript_stack_all_spls, var = "sample")
all_doses$sample <- gsub("-", "_", all_doses$sample)
all_doses <- arrange(all_doses, sample)
all_doses_mrna <- column_to_rownames(all_doses, var = "sample" )
#removing dose sort column
all_doses_mrna <- all_doses_mrna[,-13198]

all_doses_cyto <- rownames_to_column(cyto_data_stack, var = "sample")
all_doses_cyto$sample <- gsub("-", "_", all_doses_cyto$sample)
all_doses_cyto$sample <- gsub("V2", "1", all_doses_cyto$sample)
all_doses_cyto$sample <- gsub("V5", "2", all_doses_cyto$sample)
all_doses_cyto$sample <- gsub("V11", "3", all_doses_cyto$sample)
all_doses_cyto <- column_to_rownames(all_doses_cyto, var = "sample" )

#selecting significant Cytokines IP-10, MCP-2, I-TAC, MCP-1, LOX-1, IFN-gamma
all_doses_cyto_sig <- all_doses_cyto[,c(1:2, 4:5, 7,12) ]

# set a list of all the X dataframes
data = list(Transcriptomics = all_doses_mrna,
           Secretomics = all_doses_cyto_sig)
           



lapply(data, dim) # check their dimensions

#Selecting sample_id.dose and dose number 
diablo_groups <- transcript_stack_all[,c(1,13200)]

d_group1 <- as.data.table(trans_groups_alldoses)
d_group2 <- as.data.table(trans_groups_alldoses)
d_group3 <- as.data.table(trans_groups_alldoses)
d_group1$sample <-  paste(d_group1$sample, "_1",  sep = "")
d_group2$sample <-  paste( d_group2$sample,"_2", sep = "")
d_group3$sample <-  paste(d_group3$sample,"_3",  sep = "")

# d_group_pos <- d_group1[c(1,3,9:12),]
# d_group_neg <- d_group1[c(2,4,5:8),]
# d_group_pos <- bind_rows(d_group_pos, d_group2[c(1,3,9:12),]) 
# d_group_neg <- bind_rows(d_group_neg, d_group2[c(2,4,5:8),]) 
# d_group_pos <- bind_rows(d_group_pos, d_group3[c(1,3,9:12),]) 
# d_group_neg <- bind_rows(d_group_neg, d_group3[c(2,4,5:8),]) 
dia_group <- d_group1
dia_group <- bind_rows(dia_group, d_group2) 
dia_group <- bind_rows(dia_group, d_group3) 
dia_group <- arrange(dia_group, sample)
dia_group_s <- as.factor(dia_group$group)

dia_group_factor <- factor(dia_group_s, levels = c("neg", "pos"))


##FACTORS WITH TIMEPOINTS
d_group1pn <- d_group1
d_group2pn <- d_group2
d_group3pn <- d_group3

d_group1pn$group <-  paste(d_group1pn$group, "_1",  sep = "")
d_group2pn$group <-  paste( d_group2pn$group,"_2", sep = "")
d_group3pn$group <-  paste(d_group3pn$group,"_3",  sep = "")

# d_group_pos <- d_group1[c(1,3,9:12),]
# d_group_neg <- d_group1[c(2,4,5:8),]
# d_group_pos <- bind_rows(d_group_pos, d_group2[c(1,3,9:12),]) 
# d_group_neg <- bind_rows(d_group_neg, d_group2[c(2,4,5:8),]) 
# d_group_pos <- bind_rows(d_group_pos, d_group3[c(1,3,9:12),]) 
# d_group_neg <- bind_rows(d_group_neg, d_group3[c(2,4,5:8),]) 
dia_group_time <- d_group1pn
dia_group_time <- bind_rows(dia_group_time, d_group2pn) 
dia_group_time <- bind_rows(dia_group_time, d_group3pn) 
dia_group_time <- arrange(dia_group_time, sample)
dia_group_time_dose <- as.factor(dia_group_time$group)

dia_group_factor_time <- factor(dia_group_time_dose, levels = c("neg_1", "neg_2", "neg_3", "pos_1", "pos_2", "pos_3"))



```

## Setting up diablo objects
```{r}

list.keepX = c(12, 12) # select arbitrary values of features to keep
list.keepY = c(12, 12)
transcript_d1x <- transcript_d1[-8,]

#dose 1 cyto v dose 2 cyto
cyto_d1x <- cyto_d1[-8,]
cyto_d2x <- cyto_d2[,-c(11:12,20)]

# generate three pairwise PLS models
transcript_stack_all_spls <- transcript_stack_all_spls[,-13198]
pls1 <- spls(transcript_stack_all_spls, cyto_data_stack, 
             keepX = list.keepX, keepY = list.keepY) 
pls2 <- spls(transcript_d1x, transcript_d2, 
             keepX = list.keepX, keepY = list.keepY)
pls3 <- spls(cyto_d1x, cyto_d2x, 
             keepX = list.keepX, keepY = list.keepY)

# plot features of first PLS
plotVar(pls1, cutoff = 0.5, title = "(a) mRNA vs Cytokines", 
        legend = c("mRNA", "Cytokines"), 
        var.names = T, style = 'graphics', 
        pch = c(16, 17), cex = c(2,2), 
        col = c('darkorchid', 'lightgreen'))

{
  pdf(paste0("../",diablo_figures.dir,"MixOmics_PLS1_Variance_mRNAvCytokines_All_doses.pdf"),  width = 50, height = 50)
  
# plot features of first PLS
plotVar(pls1, cutoff = 0.5, title = "(a) mRNA vs Cytokines", 
        legend = c("mRNA", "Cytokines"), 
        var.names = T, style = 'graphics', 
        pch = c(16, 17), cex = c(2,2), 
        col = c('darkorchid', 'lightgreen'))

  dev.off()
}

# plot features of second PLS
plotVar(pls2, cutoff = 0.5, title = "(b) Dose 1 vs 2 Transcript", 
        legend = c("Dose 1", "Dose 2"), 
        var.names = T, style = 'graphics', 
        pch = c(16, 17), cex = c(2,2), 
        col = c('darkorchid', 'lightgreen'))
{
  pdf(paste0("../",diablo_figures.dir,"MixOmics_PLS1_Dose1v2_Transcript_All_doses.pdf"),  width = 20, height = 20)
  
# plot features of second PLS
plotVar(pls2, cutoff = 0.5, title = "(b) Dose 1 vs 2 Transcript", 
        legend = c("Dose 1", "Dose 2"), 
        var.names = T, style = 'graphics', 
        pch = c(16, 17), cex = c(2,2), 
        col = c('darkorchid', 'lightgreen'))

  dev.off()
}

# plot features of third PLS
plotVar(pls3, cutoff = 0.5, title = "(c) Dose 1 v 2 Cytokines", 
        legend = c("Dose 1", "Dose 2"), 
        var.names = T, style = 'graphics', 
        pch = c(16, 17), cex = c(2,2), 
        col = c('darkorchid', 'lightgreen'))


{
  pdf(paste0("../",diablo_figures.dir,"MixOmics_PLS1_Dose1v2_cytokines_All_doses.pdf"),  width = 20, height = 20)
  
# plot features of third PLS
plotVar(pls3, cutoff = 0.5, title = "(c) Dose 1 v 2 Cytokines", 
        legend = c("Dose 1", "Dose 2"), 
        var.names = T, style = 'graphics', 
        pch = c(16, 17), cex = c(2,2), 
        col = c('darkorchid', 'lightgreen'))

 dev.off()
}

# calculate correlation of mRNA and Cytokines
cor(pls1$variates$X, pls1$variates$Y) 

# calculate correlation of Dose 1 and 2 transcripts
cor(pls2$variates$X, pls2$variates$Y) 

# calculate correlation of Dose 1 and 2 Cytokines
cor(pls3$variates$X, pls3$variates$Y)


```

##mDIABLO Set-up and experiment

```{r}
set.seed(123) # for reproducibility, remove for normal use

X = data
#Y = dia_group_factor # used for omics data not separated by dose
Y = dia_group_factor_time
lapply(data, dim) # check their dimensions
summary(Y)
# for square matrix filled with 0.1s
design = matrix(0.1, ncol = length(data), nrow = length(data), 
                dimnames = list(names(data), names(data)))
diag(design) = 0 # set diagonal to 0s

design


# form basic DIABLO model
basic.diablo.model = block.splsda(X = data, Y = Y, ncomp = 5, design = design) 

# run component number tuning with repeated CV
perf.diablo = perf(basic.diablo.model, 
                  # validation = 'Mfold', folds = 6, nrepeat = 6,
                   validation = 'loo',  
                  signif.threshold = 0.05,
                   ) 

plot(perf.diablo) # plot output of tuning

{
 pdf(paste0("../",diablo_figures.dir,"mDIABLO_error.pdf"),  width = 10, height = 10)
plot(perf.diablo) # plot output of tuning
dev.off()
}


# set the optimal ncomp value
ncomp = perf.diablo$choice.ncomp$WeightedVote["Overall.BER", "mahalanobis.dist"] 

# show the optimal choice for ncomp for each dist metric
perf.diablo$choice.ncomp$WeightedVote 

# set grid of values for each component to test
test.keepX = list (mRNA = c(5:9, seq(10, 18, 2), seq(20,30,5)), 
                   proteomics = c(5:9, seq(10, 18, 2), seq(20,30,5)))

# run the feature selection tuning
tune.TCGA = tune.block.splsda(X = data, Y = Y, ncomp = 2, 
                              test.keepX = test.keepX, design = design,
                              #validation = 'Mfold', folds = 5, nrepeat = 1,
                              validation = 'loo',
                              signif.threshold = 0.05,
                              dist = "mahalanobis.dist"
                              #BPPARAM =
                              ) #maybe centroid.dist??

list.keepX = tune.TCGA$choice.keepX # set the optimal values of features to retain
list.keepX

# set the optimised DIABLO model
final.diablo.model = block.splsda(X = data, Y = Y, ncomp = 2, 
                          keepX = list.keepX, design = design)

final.diablo.model$design # design matrix for the final model

# the features selected to form the first component
selectVar(final.diablo.model, block = 'Transcriptomics', comp = 1)$Transcriptomics$name
selectVar(final.diablo.model, block = 'Transcriptomics', comp = 2)$Transcriptomics$name
selectVar(final.diablo.model, block = 'Secretomics', comp = 1)$Secretomics$name


```

#Diablo plots
```{r}

{
 pdf(paste0("../",diablo_figures.dir,"mDIABLO_Model_ellipse_all_doses_sig_cyto.pdf"),  width = 10, height = 10)
plotDiablo(final.diablo.model, ncomp = 1, )

dev.off()
}

{
  pdf(paste0("../",diablo_figures.dir,"Diablo.model.circos_r.65_timepoint_sig_sig_cyto NEW COLORS.pdf"),  width = 10, height = 10)
  circosPlot(final.diablo.model, cutoff = 0.65, line = T,
           color.blocks= c('darkorchid', 'lightgreen'),
           color.cor = c("chocolate3","grey20"), size.labels = 1.7, size.variables = .6 , 
           color.Y =c( "gray50", "gray30", "black", "royalblue2", "blue2", "navy" )
           #color.Y =c( "lightgray", "royalblue" )
           )
  dev.off()
}


circosPlot(final.diablo.model, cutoff = 0.7, line = T,
           color.blocks= c('darkorchid', 'lightgreen'),
           color.cor = c("chocolate3","grey20"), size.labels = 1.5,
         # showIntraLinks = T
           )

{
  pdf(paste0("../",diablo_figures.dir,"diablo_circos_all_doses_r.7_sig_cyto.pdf"),  width = 10, height = 10)
 # pdf("circos_all_doses_23_11_21.pdf")
  circosPlot(final.diablo.model, cutoff = 0.7, line = T,
           color.blocks= c('darkorchid', 'lightgreen'),
           color.cor = c("chocolate3","grey20"), size.labels = 1.5, size.variables = .5
           )
  dev.off()
}


{
  pdf(paste0("../",diablo_figures.dir,"variance_sample_all_doses_sig_cyto.pdf"),  width = 10, height = 10)
plotIndiv(final.diablo.model, ind.names = F, legend = TRUE,  color.Y =c( "gray50", "gray30", "black", "royalblue2", "blue2", "royalblue4" ) ,
          title = 'DIABLO Model Component Analysis')
  dev.off()
}

plotArrow(final.diablo.model, c(1, 2), ind.names = FALSE, legend = TRUE, 
          title = 'DIABLO Arrow Plot')

# {
#   pdf(paste0("../",diablo_figures.dir,"Arrow_Plot_all_doses.pdf"),  width = 50, height = 50)
#  
# arrow
#   dev.off()
# }


{
  pdf(paste0("../",diablo_figures.dir,"Correlation_circleplot_all_doses_sig_cyto.pdf"),  width = 15, height = 15)
plotVar(final.diablo.model, var.names = FALSE, 
        style = 'graphics', legend = TRUE,
        pch = c(16, 17), cex = c(2,2), 
        col = c('darkorchid',  'lightgreen'))
  dev.off()
} 


{
  pdf(paste0("../",diablo_figures.dir,"Network_all_doses_timepoint_2_sig_sig_cytos.pdf"),  width = 10, height = 10)
 # pdf("circos_all_doses_23_11_21.pdf")
network(final.diablo.model, blocks = c(1,2),
         color.node = c('lightblue',  'forestgreen'), cutoff = 0.4, size.node = .17)

  dev.off()
}

{
  pdf(paste0("../",diablo_figures.dir,"Contributions_component_all_doses_sig_cyto.pdf"),  width = 20, height = 20)
plotLoadings(final.diablo.model, comp = 1, contrib = 'max', method = 'median')
plotLoadings(final.diablo.model, comp = 2, contrib = 'max', method = 'median')

  dev.off()
}



{
  pdf(paste0("../",diablo_figures.dir,"DiabloMODEL_heatmap_all_doses_sig_cyto_no_patnumbers.pdf"),  width = 15, height = 15)

cimDiablo(final.diablo.model, comp = c(1,2), legend.position = "right", size.legend = 1.2, margins = c(6, 15), row.names = F,
          color.Y = c("gray50", "gray30", "black", "royalblue2", "blue2", "royalblue4" ) 
                    )

  dev.off()
}

{
  pdf(paste0("../",diablo_figures.dir,"DiabloMODEL_sample_plots_Comp_all_doses_sig_cyto.pdf"),  )

plotIndiv(final.diablo.model, ind.names = F, legend = TRUE, title = 'DIABLO sample plots comp 1 - 2', blocks = 'average', style = 'graphics', ellipse = T )

  dev.off()
}

{
  pdf(paste0("../",diablo_figures.dir,"DiabloMODEL_samplevar_plots_all_doses_sig_cyto.pdf"),  )

plotIndiv(final.diablo.model, ind.names = F, legend = TRUE,
          title = 'DIABLO Variance Plots by Variable')
  dev.off()
}


# plotIndiv(final.spls.data, group = nutrimouse$genotype,
#           rep.space = 'XY-variate',
#           ellipse = TRUE,  # plot using the ellipses
#           legend = TRUE)


{
  pdf(paste0("../",result.dir,"DiabloMODEL_all_doses_sig_cyto.pdf"),  )

plotIndiv(final.diablo.model)

  dev.off()
}



#cimDiablo(final.diablo.model)


{
  pdf(paste0("../",result.dir,"DiabloMODEL_modelblock_all_doses_sig_cyto.pdf"),  )

cimDiablo(final.diablo.model, comp = c(1,2), legend.position = "right", size.legend = 1.1, margins = c(6, 15))

  dev.off()
}



```

##  Plots showing expression values Fold change of selected genes across time points and exploration of correlation between genes and cytokines of interest
```{r}

# Fold Change Expression of GPR84 across all doses in both groups
trans_groups2 <- trans_groups[,-c(4,7:8)]
trans_groups2 <- bind_rows(trans_groups_alldoses, trans_groups_alldoses, trans_groups_alldoses)

fc_plots$group <- trans_groups2$group
fc_plots <- bind_cols(transcript_stack_all, trans_groups2$group)

ggplot(fc_plots,
  aes(x = dose, y = GPR84, color =  group, group = group)) + 
  geom_point() + stat_summary(fun.y=mean, geom="line") +
  #scale_y_log10()  +
  ggtitle("GPR84") +
  labs(x = "Dose", y = "Transcripts per million FC") +
  scale_color_manual(values = c("#A9A9A9", "#A9CAF6")) +
  #theme_prism(base_line_size = .5, base_fontface = "plain", axis_text_angle = 45) +
  theme(axis.ticks = element_line(size = .5), legend.position = "right",
        aspect.ratio = 1,
        legend.title = element_text())


ggsave(filename = paste0("../", result.dir,"GPR84.pdf"),
      height = 5, width = 5)


## Correlation between MCP-1 & IFIT3
mcp1_data <- cyto_ds[,c(1:3,24)]
mcp1_data <- mcp1_data %>% dplyr::filter(Timepoint %in% time)


ifit3_data <- dplyr::select(trans_ds, contains('IFIT3'))

ifit3_data <- bind_cols(trans_groups,ifit3_data)

ifit3_data_long <- pivot_longer(ifit3_data, cols = !sample:group, values_to = "IFIT3")

df_scatter_ifit3 <- bind_cols(mcp1_data, ifit3_data_long[,4])

ggplot(df_scatter_ifit3, aes(x = IFIT3, y = df_scatter_ifit3$`MCP-1`, colour = group )) +
  geom_point(aes(shape = group,) ) + 
  geom_smooth(aes( colour = group), method = "lm", se = F, ) + 
  stat_cor(aes(group = group), method = "spearman" ) +
  labs(title = "Fold Change Scatter Plot", x = "IFIT3", y = "MCP-1") +
  theme_minimal() 

```

## SessionInfo
```{r}
sessionInfo()
```


