---
title: "Orebro RNA-seq analysis"
author: "Rodrigo Arcoverde Cerveira & Gustav Joas"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output: html_document
abstract: Örebro study RNA-seq data analysis.
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/lab_book/")})
editor_options: 
  chunk_output_type: inline
---

```{r, global.options, include=FALSE}
# Set up figures and chunks messages

knitr::opts_knit$set(
  echo = TRUE,
  root.dir = getwd(),
  fig.width = 6, fig.height = 5,
  fig.align = "center",
  out.width = 768,
  fig.pos = "H",
  warning = FALSE,
  message = FALSE
)
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.width = 6, fig.height = 5,
  fig.align = "center",
  out.width = 768,
  fig.pos = "H"
)

result.dir <- paste0("results/", Sys.Date(), "/")
figures.dir <- paste0("results/", Sys.Date(), "/", "figures/")

# Create result.dir with date in if not existent
ifelse(isFALSE(dir.exists(paste0("../", result.dir))),
  dir.create(paste0("../", result.dir), recursive = TRUE),
  "Result directory for today exists already!"
)

ifelse(isFALSE(dir.exists(paste0("../", figures.dir))),
  dir.create(paste0("../", figures.dir), recursive = TRUE),
  "Result directory for today exists already!"
)

options(stringsAsFactors = FALSE)
```

## Loading libraries

```{r, results='hide'}
library(data.table)
library(kableExtra)
library(forcats)
library(mgsub)
library(DESeq2)
library(pheatmap)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(limma)
library(edgeR)
library(enrichR)
library(gridExtra)
library(stringr)
library(ggVennDiagram)
library(dplyr)
library(ensembldb)
library(EnsDb.Hsapiens.v86)
library(factoextra)
library(rafalib)
library(plotly)
library(tibble)
library(devtools)
library(clusterProfiler)
library(enrichplot)
library(pathview)
library(ggprism)
```

## Load and wrangle raw counts

```{r, results='hide'}
# Load batch 1 raw counts 
counts_raw_1 <- data.table::fread("../data/processed_data_nfcore/salmon/salmon.merged.gene_counts.tsv",
                                  header = TRUE, sep = "\t") %>% 
  as.data.frame() %>%
  select(-gene_id)

counts_raw_1_aggr <- aggregate(counts_raw_1[, -1], 
                        list(gene_name = counts_raw_1[, 1]), 
                        FUN = sum)

#load batch 3 raw counts
counts_raw_3 <- data.table::fread("../data/processed_data_nfcore/batch_3/salmon.merged.gene_counts.tsv",
                                  header=TRUE, sep="\t") %>% 
 as.data.frame() %>%
select(-gene_id)

counts_raw_3_aggr <- aggregate(counts_raw_3[, -1], 
                        list(gene_name = counts_raw_3[, 1]), 
                        FUN = sum)

#check rownames
identical(rownames(counts_raw_3_aggr), rownames(counts_raw_1_aggr))

#merge aggregated dataframes
counts_raw <- left_join(counts_raw_1_aggr, counts_raw_3_aggr, by="gene_name")

colnames(counts_raw) <- sub("_L001|_L002", "",colnames(counts_raw))


# Aggregate and sum up genes with same gene symbol (basically non-coding RNAs or ribosomic)
counts_raw_aggr <- aggregate(counts_raw[, -1], 
                        list(gene_name = counts_raw[, 1]), 
                        FUN = sum)


#Change row names to gene names
rownames(counts_raw) <- counts_raw$gene_name
counts_raw <- dplyr::select(counts_raw, -c(gene_name))
```

## Load metadata and match with raw counts

```{r, results='hide'}
# Load sample metadata
sampleTable <- read.table("../data/metadata/sample_metadata_2.csv", 
                          sep = ",", header = TRUE) %>%
  dplyr::select(-c(age, sex))

# Load patient metadata
patient_metadata <- read.table("../data/metadata/patient_metadata.csv", 
                               sep = ",", header = TRUE)

# Join sample and patient metadata
sampleTable <- sampleTable %>% left_join(patient_metadata, by = "subject_ID")

# Change visits to dose and time point...
sampleTable <- sampleTable %>%
  mutate(
    dose = as.factor(plyr::mapvalues(visit,
      from = c(1, 2, 4, 5, 10, 11, 12),
      to   = c("0_dose1", "24_dose1", "0_dose2", "24_dose2", "0_dose3", "24_dose3", "48_dose3")
    )),
    group = as.factor(group)
  ) %>%
  # ...(sex) 0 and 1 to male and female...
  mutate(
    sex = as.factor(plyr::mapvalues(sex,
      from = c(0, 1),
      to   = c("male", "female")
    )),
    group = as.factor(sex)
  ) %>%
  # ...(prior covid) 0 and 1 to neg and pos
  mutate(
    prior_covid19 = as.factor(plyr::mapvalues(prior_covid19,
      from = c(0, 1),
      to   = c("neg", "pos")
    )),
    group = as.factor(prior_covid19)
  )

# Remove 24dose_2 from subject 21 (missing 0dose_2)
sampleTable <- subset(sampleTable, sample_ID != "P22761_1037_S37")

# Remove two samples from subject 7 (sample replicate)
sampleTable <- subset(sampleTable, sample_ID != "P26208_1060_S62")
sampleTable <- subset(sampleTable, sample_ID != "P27910_1057_S1")

# Remove corresponding samples from count table
counts_raw <- subset(counts_raw, select = -c(P27910_1057_S1,P26208_1060_S62,P22761_1037_S37))

# Match count data with sample table
colnames(counts_raw) <- sampleTable$sample_ID
counts_raw <- counts_raw[, pmatch(sampleTable$sample_ID, colnames(counts_raw))]
all(rownames(sampleTable$sample_ID) == colnames(counts_raw))
```

## Basic Quality Control

### Inspect raw data

Plotting raw counts for first visual inspection. The boxplot shows median values of zero across all samples. In the histogram there is a huge peak of zero counts. Both indicate that the data set would benefit from a low count filtering.

```{r}
# Visualize distribution of raw counts with boxplot and density plot
boxplot(log2(as.matrix(counts_raw) + 1),
  ylab = expression("Log"[2] ~ "Read counts"),
  las = 2,
  main = "Raw data"
)
hist(log2(as.matrix(counts_raw) + 1),
  ylab = "",
  las = 2,
  main = "Raw data"
)

```

```{r, echo=FALSE, results='hide'}
# Saving plots
{
  pdf(paste0("../", figures.dir, "raw_counts_QC.pdf"), 
      width = 10, height = 8, compress = TRUE)
  rafalib::mypar(1, 2, mar = c(10, 3, 3, 2))
  boxplot(outline = FALSE,log2(as.matrix(counts_raw) + 1),
    ylab = expression("Log"[2] ~ "Read counts"),
    las = 2,
    main = "Raw data"
  )
  hist(log2(as.matrix(counts_raw) + 1),
    ylab = "",
    las = 2,
    main = "Raw data"
  )
  par(mfrow = c(1, 1))
  garbage <- dev.off()
}
```

### Filtering

Plot number of genes detected across samples. All samples are more or less close to average so we don't need to discard any samples.

```{r}
barplot(colSums(counts_raw > 3),
  ylab = "Number of detected genes",
  las = 2
) +
  abline(h = median(colSums(counts_raw > 3)))
```

```{r, echo=FALSE, results='hide'}
# Saving plots
{
  pdf(paste0("../", figures.dir, "number_detected_genes.pdf"), 
      width = 10, height = 8, compress = TRUE)
  par(mar = c(10, 4, 3, 4))
  barplot(colSums(counts_raw > 3),
    ylab = "Number of detected genes",
    las = 2
  ) +
    abline(h = median(colSums(counts_raw > 3)))
  dev.off()
}
```

Removing reads with the log2 of the counts per million (cpm) lower than 1 (= 45896 genes). Plot genes detection rate across samples, comparing raw and filtered counts.

```{r}
meanLog2CPM <- rowMeans(log2(cpm(counts_raw) + 1))
counts_filtered <- counts_raw[meanLog2CPM > 1, ]

hist(rowSums(counts_raw > 3), main = title("Raw Counts"))
hist(rowSums(counts_filtered > 3), main = title("Filtered Counts"))
```

```{r, echo=FALSE, results='hide'}
# Saving plots
{
  pdf(paste0("../", figures.dir, "detection_rate_raw_filtered.pdf"), 
      width = 10, height = 8, compress = TRUE)
  par(mar = c(10, 4, 3, 4))
  hist(rowSums(counts_raw > 3), main = title("Raw Counts"))
  hist(rowSums(counts_filtered > 3), main = title("Filtered Counts"))
  par(mfrow = c(1, 1))
  dev.off()
}
```

Plot distribution of the filtered counts

```{r}
boxplot(log2(as.matrix(counts_filtered) + 1),
  ylab = expression("Log"[2] ~ "Read counts"),
  las = 2,
  main = "Filtered data"
)
hist(log2(as.matrix(counts_filtered) + 1),
  ylab = "",
  las = 2,
  main = "Filtered data"
)
```

```{r, echo=FALSE, results='hide'}
{
  pdf(paste0("../", figures.dir, "filtered_counts_distribution.pdf"), 
      width = 10, height = 8, compress = TRUE)
  rafalib::mypar(1, 2, mar = c(10, 3, 3, 2))
  boxplot(log2(as.matrix(counts_filtered) + 1),
    ylab = expression("Log"[2] ~ "Read counts"),
    las = 2,
    main = "Filtered data"
  )
  hist(log2(as.matrix(counts_filtered) + 1),
    ylab = "",
    las = 2,
    main = "Filtered data"
  )
  par(mfrow = c(1, 1))
  dev.off()
}
```

### DESeq object creation

Generating three separate DESeq objects with different design parameters to allow for more comparisons. Controlling for age and sex.

```{r, results='hide'}
#Remove vaccine/sars-cov-2 related genes

# Prepare for DESeq by creating new variables and factorizing 
sampleTable <- sampleTable %>%
  mutate(
    conditions = paste0(sampleTable$group, "_", sampleTable$dose),
    time = gsub(".{6}$", "", conditions),
    conditions = factor(conditions),
    time = factor(time),
    sex = factor(sex),
    age = factor(age),
    dose = factor(dose),
    group = factor(group),
    batch = factor(batch)
  )

# create 3 DESeq objects comparing against conditions, time and dose
dds1 <- DESeqDataSetFromMatrix(
  countData = round(counts_filtered),
  colData = as.data.frame(sampleTable),
  design = ~ sex + age + batch + conditions
)

dds2 <- DESeqDataSetFromMatrix(
  countData = round(counts_filtered),
  colData = as.data.frame(sampleTable),
  design = ~ sex + age + batch + time
)

dds3 <- DESeqDataSetFromMatrix(
  countData = round(counts_filtered),
  colData = as.data.frame(sampleTable),
  design = ~ sex + age + batch + dose
)
```

### Normalization and Quality Control

Normalize with variance stabilizing transformation Plot distribution of data after normalization

```{r}
# Normalize with variance stabilizing transformation for later PCA and heatmap
counts_vst_normalized <- vst(dds1, blind = TRUE)

vst_matrix <- assay(counts_vst_normalized)
hist(vst_matrix)
boxplot(vst_matrix, 
        ylab = expression("Log"[2] ~ "Read counts"), 
        las = 2, 
        main = "VST")
```

```{r, echo=FALSE, results='hide'}
# Saving plots

{
  pdf(paste0("../", figures.dir, "VST.pdf"), 
      width = 10, height = 8, compress = TRUE)
  rafalib::mypar(1, 2, mar = c(10, 3, 3, 2))
  hist(vst_matrix)
  boxplot(vst_matrix, 
        ylab = expression("Log"[2] ~ "Read counts"), 
        las = 2, 
        main = "VST")
  par(mfrow = c(1, 1))
  dev.off()
}
```

### Heatmap

```{r}
# Sample heatmap
sampleDist <- cor(vst_matrix, method = "spearman")

Metadata <- data.frame(sampleTable$group, sampleTable$dose, sampleTable$batch)
names(Metadata) <- c("Group", "Dose", "batch")
rownames(Metadata) <- sampleTable$sample_ID

# Plot heatmap
colors <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(255)

Heatmap <- pheatmap(sampleDist,
  color = colors,
  clustering_distance_rows = as.dist(1 - sampleDist),
  clustering_distance_cols = as.dist(1 - sampleDist),
  show_rownames = FALSE,
  show_colnames = FALSE,
  clustering_method = "ward.D2",
  annotation_col = Metadata
)

print(Heatmap)
```

```{r, echo=FALSE}
# Saving heatmap
   ggsave(
      filename = paste0("../", figures.dir,"Heatmap.png"),
      height = 8, width = 10,
      plot = Heatmap)
```

### Principal Component Analysis (PCA)

Dimensionality reduction for evaluating outliers and global sample clusters for both quality control but also for primary exploratory analysis.

```{r}
# PCA plot
pcaRes <- prcomp(t(assay(counts_vst_normalized)))
varExp <- round(summary(pcaRes)[[1]], digits = 2)

pcaDF <- data.frame(
  PC1 = pcaRes$x[, 1],
  PC2 = pcaRes$x[, 2],
  Group = sampleTable$group,
  Sample = sampleTable$sample_ID,
  Dose = sampleTable$dose,
  Batch = sampleTable$batch
)

pcaPlot <- ggplot(pcaDF, mapping = aes(x = PC1, y = PC2, color = Batch, label = Sample)) +
  geom_point(aes(shape = Group), size = 3) +
  labs(x = paste0("PC1 (", varExp[1], " %)"), 
       y = paste0("PC2 (", varExp[2], " %)")) +
  theme(axis.text = element_text(size = 12), 
        legend.text = element_text(size = 10)) +
  scale_color_manual(values = brewer.pal(3, "Accent")) +
  cowplot::theme_cowplot()

print(pcaPlot)

# Scree plot, showing contribution of principal components in descending order
pca.var <- pcaRes$sdev^2
pca.var.per <- round(pca.var / sum(pca.var) * 100, 1)

barplot(pca.var.per, main = "Scree Plot", 
        xlab = "Principal Component", 
        ylab = "Percent Variation")

# Explore which variables contribute the most for each PC
var <- get_pca_var(pcaRes)
var$contrib %>%
  as.data.frame() %>%
  arrange(desc(Dim.1)) %>%
  head() %>%
  kableExtra::kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r, echo=FALSE, results='hide'}
# Saving PCA plot
   ggsave(
      filename = paste0("../", figures.dir,"PCA.pdf"),
      height = 8, width = 10,
      plot = pcaPlot)

# Saving Scree plot
{
  pdf(paste0("../", figures.dir, "Scree_plot.pdf"), 
      width = 10, height = 8, compress = TRUE)
  barplot(pca.var.per,
    main = "Scree Plot",
    xlab = "Principal Component",
    ylab = "Percent Variation"
  )
  par(mfrow = c(1, 1))
  dev.off()
}
```


## Differential Gene Expression Analysis

Differential Gene Expression Analysis using DESeq2. Comparing parameters; dose, time point, prior infection status.

```{r, results='hide'}
# Run the DESeq2 analysis
dds1 <- DESeq(dds1)
dds2 <- DESeq(dds2)
dds3 <- DESeq(dds3)

# Contrast
resultsNames(dds1)
resultsNames(dds2)
resultsNames(dds3)

# Create function for makings comparisons with default being conditions and dds1
compare_DEGs <- function(variable1, variable2, condition = "conditions", dds = dds1) {
  as.data.frame(results(dds, contrast = c(condition, variable1, variable2)))
}

# Baselines DEGs
base_pos_neg_d1 <- compare_DEGs("pos_0_dose1", "neg_0_dose1")
base_pos_neg_d2 <- compare_DEGs("pos_0_dose2", "neg_0_dose2")
base_pos_neg_d3 <- compare_DEGs("pos_0_dose3", "neg_0_dose3")

base_pos_d1_d2 <- compare_DEGs("pos_0_dose2", "pos_0_dose1")
base_pos_d1_d3 <- compare_DEGs("pos_0_dose3", "pos_0_dose1")
base_pos_d2_d3 <- compare_DEGs("pos_0_dose2", "pos_0_dose3")

base_neg_d1_d2 <- compare_DEGs("neg_0_dose2", "neg_0_dose1")
base_neg_d1_d3 <- compare_DEGs("neg_0_dose3", "neg_0_dose1")
base_neg_d2_d3 <- compare_DEGs("neg_0_dose2", "neg_0_dose3")

# 0-24h and 0-48 (3rd dose) DEGs
neg_d1 <- compare_DEGs("neg_24_dose1", "neg_0_dose1")
neg_d2 <- compare_DEGs("neg_24_dose2", "neg_0_dose2")
neg_d3_24 <- compare_DEGs("neg_24_dose3", "neg_0_dose3")
neg_d3_48 <- compare_DEGs("neg_48_dose3", "neg_0_dose3")

pos_d1 <- compare_DEGs("pos_24_dose1", "pos_0_dose1")
pos_d2 <- compare_DEGs("pos_24_dose2", "pos_0_dose2")
pos_d3_24 <- compare_DEGs("pos_24_dose3", "pos_0_dose3")
pos_d3_48 <- compare_DEGs("pos_48_dose3", "pos_0_dose3")

# Results dds2
pos <- compare_DEGs("pos_24", "pos_0", condition = "time", dds = dds2)
neg <- compare_DEGs("neg_24", "neg_0", condition = "time", dds = dds2)

# Results dds3
d1 <- compare_DEGs("24_dose1", "0_dose1", condition = "dose", dds = dds3)
d2 <- compare_DEGs("24_dose2", "0_dose2", condition = "dose", dds = dds3)
d3_24 <- compare_DEGs("24_dose3", "0_dose3", condition = "dose", dds = dds3)
d3_48 <- compare_DEGs("48_dose3", "0_dose3", condition = "dose", dds = dds3)
```

### Volcano plots


Volcano plot for data exploratory analysis, filtering for significant values with False Discovery Rate < 0.05 and log fold change greater than 1.

```{r, fig.width=18, fig.height=12, fig.keep=10:17}
results_list <- list(base_pos_neg_d1,
                    base_pos_neg_d2,
                    base_pos_neg_d3, 
                    base_pos_d1_d2, 
                    base_pos_d1_d3, 
                    base_pos_d2_d3, 
                    base_neg_d1_d2, 
                    base_neg_d1_d3, 
                    base_neg_d2_d3, 
                    neg_d1,
                    neg_d2, 
                    neg_d3_24, 
                    neg_d3_48, 
                    pos_d1, 
                    pos_d2, 
                    pos_d3_24, 
                    pos_d3_48, 
                    pos, 
                    neg, 
                    d1, 
                    d2, 
                    d3_24, 
                    d3_48)
                     
names(results_list) <- c("Baseline Dose 1 Conv vs Naïve", 
                         "Baseline Dose 2 Conv vs Naïve", 
                         "Baseline Dose 3 Conv vs Naïve", 
                         "Baseline Conv Dose 1 and 2", 
                         "Baseline Conv Dose 1 and 3", 
                         "Baseline Conv Dose 2 and 3", 
                         "Baseline Naïve Dose 1 and 2", 
                         "Baseline Naïve Dose 1 and 3", 
                         "Baseline Naïve Dose 2 and 3",
                         "Naïve Dose 1 0-24 hours", 
                         "Naïve Dose 2 0-24 hours", 
                         "Naïve Dose 3 0-24 hours", 
                         "Naïve Dose 3 0-48 hours", 
                         "Conv Dose 1 0-24 hours", 
                         "Conv Dose 2 0-24 hours", 
                         "Conv Dose 3 0-24 hours", 
                         "Conv Dose 3 0-48 hours", 
                         "Conv 0-24 hours",
                         "Naïve 0-24 hours", 
                         "Dose 1 0-24 hours", 
                         "Dose 2 0-24 hours",
                         "Dose 3 0-24 hours",
                         "Dose 3 0-48 hours") 

    results_list_plot <- lapply(results_list, function(x) {
  x <- x %>% mutate(test_padj = case_when(padj < 0.05 & log2FoldChange >= 1 ~ "Up regulated",
                                     padj < 0.05 & log2FoldChange <= -1 ~ "Down regulated",
                                     TRUE ~ "Not significant"),
                    test_pvalue = case_when(pvalue < 0.05 & log2FoldChange >= 1 ~ "Up regulated",
                                     pvalue < 0.05 & log2FoldChange <= -1 ~ "Down regulated",
                                     TRUE ~ "Not significant"),
                    )
  })
    
# function to plot volcano plots
volcano_plot <-  function(x){
  subsetted <- subset(results_list_plot[[x]] %>% tibble::rownames_to_column("gene_symbol"), abs(log2FoldChange) >= 1 & padj < 0.05)
  label_list <- c("RSAD2", "IFIT1", "CXCL10", "CMPK2", "GBP1P1", "IFI44L", "BATF2", "SIGLEC1", "IFI44", "NDUFS4", "MX1", "CXCL9", "FCGR1A", "FCGR1B", "CCL4")
  label<- subsetted %>% filter(gene_symbol %in% label_list)

  # count up regulated DEGs
  deg_number_up <- subset(
    results_list_plot[[x]] %>%
      tibble::rownames_to_column("gene_symbol"),
    log2FoldChange >= 1 & padj < 0.05
  ) %>%
    nrow()
  # count down regulated DEGs
  deg_number_down <- subset(
    results_list_plot[[x]] %>%
      tibble::rownames_to_column("gene_symbol"),
    log2FoldChange <= -1 & padj < 0.05
  ) %>%
    nrow()

  
  results_list_plot[[x]] %>% 
  ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(colour = test_padj), size = 3, alpha = 0.3) +
  scale_color_manual(values = c("Down regulated"="blue", "Not significant" ="grey80", "Up regulated"="red")) +
  xlab(expression("Fold Change (Log"[2]*")")) +
  ylab(expression("-Log"[10]*"(p-value)")) +
  labs(x = NULL, y = NULL) +
  geom_vline(xintercept = c(-1), linetype = "dotted", size = 1) + 
  geom_vline(xintercept = c(1), linetype = "dotted", size = 1) + 
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", size = 1) +
 geom_label_repel(data = label, aes(log2FoldChange,-log10(pvalue),label= gene_symbol)) +
 xlim(-2,6) +
 ylim(0,20) +
  ggtitle(x) +  
  annotate(geom='text', colour = "red", size = 10, x= 4, y=20, hjust=0, label=paste0(deg_number_up)) +
  annotate(geom='text', colour = "blue", size = 10, x= -2, y= 20, hjust=0, label=paste0(deg_number_down)) +
 ggprism::theme_prism() +
  theme(axis.ticks = element_line(size = .5),legend.position = 'none' ) 
  }


volcano_plots <- lapply(names(results_list_plot), volcano_plot)
names(volcano_plots) <- names(results_list_plot)
volcano_plots
```


```{r, echo=FALSE, results='hide'}
# save volcano plots
lapply(names(volcano_plots), function(x) {
  ggsave(
    filename = paste0("../", figures.dir, x, "_volcano_plot.pdf"), 
    width = 7, height = 7,
    plot = volcano_plots[[x]]
  )
})

```
### Enrichment analysis

Selecting two different databases for search for enrichment analysis. The databases selected were "GO_Biological_Process_2021" and "KEGG_2021_Human".

Also testing the Blood Transcriptome modules presented in [Li et al. 2014](10.1038/ni.2789) using some GSEA package.

```{r, results='hide'}
# Enrichment analysis
listEnrichrDbs()

enrichment_list_up <- lapply(results_list_plot, function(x) {
  enrichr(
    genes = rownames(x[x$test_padj == "Up regulated", ]),
    databases = c("GO_Biological_Process_2021", 
                  "KEGG_2021_Human", 
                  "GO_Molecular_Function_2021",
                  "Reactome_2022")
  )
})


enrichment_list_down <- lapply(results_list_plot, function(x) {
  enrichr(
    genes = rownames(x[x$test_padj == "Down regulated", ]),
    databases = c("GO_Biological_Process_2021", 
                  "KEGG_2021_Human", 
                  "GO_Molecular_Function_2021",
                  "Reactome_2022")
  )
})

gmt <- read.gmt("../data/blood_transcription_module/BTM_for_GSEA_20131008.gmt")
C7.gmt <- read.gmt("../data/blood_transcription_module/C7_vax.gmt")
#gmt.hallmark <- read.gmt("../data/metabolic_transcriptionmodule/MgSig_Hallmark.gmt")
#gmt.reactome <- read.gmt("../data/metabolic_transcriptionmodule/MSigDB_reactome.gmt")
```

#### Over-Representation Analysis

Over-representation looks for biological pathways that are enriched more than would be expected by chance. It only takes the genes that were found to be differentially expressed into account.

```{r echo=TRUE, fig.height=8, fig.width=15, message=FALSE, results='hide'}
# Remove items without significant results from list
enrichment_list_up[sapply(enrichment_list_up, is.null)] <- NULL
enrichment_list_down[sapply(enrichment_list_down, is.null)] <- NULL

# Optional: uncomment two lines below to remove all comparisons except naive vs conv at each timepoint
# enrichment_list_up <- enrichment_list_up[3:9]
# enrichment_list_down <- enrichment_list_down[3:6]

# function for plotting over-representation analysis
plot_enrichment <- function(enrichment_list) {
  plots_enrich <- list()
  for (pathway in 1:length(enrichment_list[[1]])) {
    rbinded_enrich_list <- lapply(enrichment_list, function(x) x[[pathway]])
    print(names(enrichment_list[[1]])[[pathway]])
    rbinded_enrich_df <- data.table::rbindlist(l = rbinded_enrich_list, idcol = TRUE, fill = TRUE)

    plots_enrich[[pathway]] <- rbinded_enrich_df %>%
      group_by(.id) %>%
      arrange(Adjusted.P.value) %>%
      slice(1:10)
  }
  plots <- lapply(plots_enrich, function(x) {
    x %>%
      ggplot(aes(x = Term, y = -log10(Adjusted.P.value))) +
      geom_bar(stat = "identity", width = 0.05) +
      geom_point(size = 3) +
      theme_minimal() +
      theme(
        text = element_text(size = 10),
        plot.title = element_text(hjust = (5 / nchar(results_list)) * 2),
        plot.margin = margin(t = 5, r = 50, b = 5, unit = "pt"),
        axis.text.y = element_text(size = 8)
      ) +
      coord_flip() +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray") +
      theme_bw() +
      labs(x = NULL, y = expression("Adjusted P value, Log"[10] * "")) +
      facet_wrap(~.id, ncol = 4)
  })
  return(plots)
}

# Plotting over-representation analysis
enrichment_plot_up <- plot_enrichment(enrichment_list_up)
enrichment_plot_down <- plot_enrichment(enrichment_list_down)

names(enrichment_plot_up) <- c(
  "GO_Biological_Process_2021_up",
  "KEGG_2021_Human_up",
  "GO_Molecular_Function_2021_up",
  "Reactome_2022_up"
)

names(enrichment_plot_down) <- c(
  "GO_Biological_Process_2021_down",
  "KEGG_2021_Human_down",
  "GO_Molecular_Function_2021_down",
  "Reactome_2022_down"
)
print(enrichment_plot_up)
print(enrichment_plot_down)
```

```{r, echo=FALSE, results='hide'}
# function for saving op plots
save_op_plots <- function(ea_plot_list) {
  lapply(names(ea_plot_list), function(x) {
    ggsave(
      filename = paste0("../", figures.dir, x, "_EA.png"),
      height = 24, width = 16,
      plot = ea_plot_list[[x]]
    )
  })
}

# saving enrichment plots to figure folder
save_op_plots(enrichment_plot_up)
save_op_plots(enrichment_plot_down)
```

### Gene Set Enrichment Analysis

GSEA looks for patterns of enriched genes which are related to each other. Compared with over-representation analysis this does not rely solely on DEGs. Therefore, this method is better when the differences in gene expression is smaller.

#### Annotations

```{r, results='hide'}
# Set desired organim to homo sapiens
organism <- "org.Hs.eg.db"
# BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
```

#### Prepare Input

```{r}
process_to_gsea <- function(x) {
  original_gene_list <- x$log2FoldChange
  names(original_gene_list) <- rownames(x)
  gene_list <- na.omit(original_gene_list)
  gene_list <- sort(gene_list, decreasing = TRUE)
}

ls_processed <- lapply(results_list, process_to_gsea)
```

#### Gene Set Enrichment Li et al BTMs

```{r, include=FALSE}
list_gse <- lapply(ls_processed, function(x) {
  GSEA(
    geneList = x,
    TERM2GENE = gmt,
    minGSSize = 3,
    maxGSSize = 800,
    nPermSimple = 10000,
    pvalueCutoff = 0.05,
    verbose = TRUE,
    pAdjustMethod = "fdr"
  )
})
```

#### Gene Set Enrichment MSigDB C7 gene sets
```{r, include=FALSE}
list_gse_C7 <- lapply(ls_processed, function(x) {
  GSEA(
    geneList = x,
    TERM2GENE = C7.gmt,
    minGSSize = 3,
    maxGSSize = 800,
    nPermSimple = 10000,
    pvalueCutoff = 0.05,
    verbose = TRUE,
    pAdjustMethod = "fdr"
  )
})
```


#### Ridgeplots BTMs

Helpful to interpret up/down-regulated pathways.

```{r, fig.width=18, fig.height=12, fig.keep=10:17}
#Optional: uncomment line below to remove all comparisons except naive vs conv at each timepoint
#list_gse_hallmark <- list_gse_hallmark[10:17]


ridgeplotting <- function(x) {
  list_gse[[x]] %>%
    ridgeplot(fill = "enrichmentScore") +
    viridis::scale_fill_viridis(option = "turbo") +
    ggtitle(x) +
    cowplot::theme_cowplot(
      font_size = 6,
      line_size = 1.5,
      rel_small = 1.6,
      rel_tiny = 1,
      font_family = "Arial",
      rel_large = 4
    ) +
    theme(
      legend.position = "right",
      legend.direction = "vertical",
      legend.key.size = unit(1, "cm"),
      legend.title = element_text(size = 16, face = "bold", family = "Arial"),
      legend.text = element_text(size = 14),
      axis.title = element_text(size = 35),
      axis.text.x = element_text(size = 16),
      axis.text.y = element_text(size = 16)
    ) +
    scale_y_discrete()
}

ridgeplots <- lapply(names(list_gse), ridgeplotting)
names(ridgeplots) <- names(list_gse)
print(ridgeplots)

```

```{r, echo=FALSE, results='hide'}
# save ridge plots
lapply(names(ridgeplots), function(x) {
  ggsave(
    filename = paste0("../", figures.dir, x, "_Reactome_plot.png"),
    plot = ridgeplots[[x]], width = 16, height = 12
  )
})
```

#### Combined GSEA heatmap

Using the Blood Transcriptome modules presented in [Li et al. 2014](10.1038/ni.2789).

```{r, gsea_heatmap}
my_palette <- colorRampPalette(c("#265891", "white", "#B80F20"))(n = 100)
sel_list_gse <- list_gse[c(
  "Naïve Dose 1 0-24 hours",
  "Naïve Dose 2 0-24 hours",
  "Naïve Dose 3 0-24 hours",
  "Naïve Dose 3 0-48 hours",
  "Conv Dose 1 0-24 hours",
  "Conv Dose 2 0-24 hours",
  "Conv Dose 3 0-24 hours",
  "Conv Dose 3 0-48 hours"
)]

combined_gsea <- data.table::rbindlist(lapply(sel_list_gse, as.data.frame), idcol = TRUE) %>%
  group_by(.id) %>%
  arrange(desc(NES)) %>%
  ungroup()

to_heatmap <- combined_gsea %>%
  select(.id, ID, NES) %>%
  filter(NES > 2.5 | NES < -2.5, !grepl("TBA", ID)) %>%
  tidyr::pivot_wider(values_from = NES, names_from = ID) %>%
  mutate(across(where(anyNA), ~ tidyr::replace_na(., 0))) %>%
  t()
colnames(to_heatmap) <- to_heatmap[1, ]
to_heatmap <- to_heatmap[-1, ]

col_order <- c(
  "Naïve Dose 1 0-24 hours",
  "Conv Dose 1 0-24 hours",
  "Naïve Dose 2 0-24 hours",
  "Conv Dose 2 0-24 hours",
  "Naïve Dose 3 0-24 hours",
  "Conv Dose 3 0-24 hours",
  "Naïve Dose 3 0-48 hours",
  "Conv Dose 3 0-48 hours"
)

to_heatmap <- to_heatmap[, col_order]

BTM_heatmap <- matrix(as.numeric(to_heatmap), ncol = ncol(to_heatmap)) %>%
    pheatmap::pheatmap(
      width = 6,
      height = 20,
      cellwidth = 20,
      cellheight = 5,
      cluster_rows = TRUE,
      cluster_cols = FALSE,
      clustering_method = "complete",
      angle_col = 45,
      border_color = "black",
      cutree_cols = 4,
      color = my_palette,
      labels_col = colnames(to_heatmap),
      labels_row = rownames(to_heatmap),
      treeheight_col = 1,
      fontsize_row = 6,
      fontsize_col = 8,
      gaps_col = c(2, 4, 6)
    )

```

```{r, echo=FALSE, results='hide'}
# save BTM heatmap
   ggsave(
      filename = paste0("../", figures.dir,"BTM_heatmap.pdf"),
      height = 8, width = 10,
      plot = BTM_heatmap)
```

##C7 Ridgeplots
```{r, fig.width=18, fig.height=12, fig.keep=10:17}
#Optional: uncomment line below to remove all comparisons except naive vs conv at each timepoint
#list_gse <- list_gse_hallmark[10:17]


ridgeplotting <- function(x) {
  list_gse_C7[[x]] %>%
    ridgeplot(fill = "enrichmentScore") +
    viridis::scale_fill_viridis(option = "turbo") +
    ggtitle(x) +
    cowplot::theme_cowplot(
      font_size = 6,
      line_size = 1.5,
      rel_small = 1.6,
      rel_tiny = 1,
      font_family = "Arial",
      rel_large = 4
    ) +
    theme(
      legend.position = "right",
      legend.direction = "vertical",
      legend.key.size = unit(1, "cm"),
      legend.title = element_text(size = 16, face = "bold", family = "Arial"),
      legend.text = element_text(size = 14),
      axis.title = element_text(size = 35),
      axis.text.x = element_text(size = 16),
      axis.text.y = element_text(size = 16)
    ) +
    scale_y_discrete()
}

ridgeplots_C7 <- lapply(names(list_gse), ridgeplotting)
names(ridgeplots_C7) <- names(list_gse_C7)
print(ridgeplots_C7)

```

#### Combined GSEA heatmap

Using the MgSigDB C7 modules 

```{r, gsea_heatmap}
my_palette <- colorRampPalette(c("#265891", "white", "#B80F20"))(n = 20)
sel_list_gse_C7 <- list_gse_C7[c(
  "Naïve Dose 1 0-24 hours",
  "Naïve Dose 2 0-24 hours",
  "Naïve Dose 3 0-24 hours",
  "Naïve Dose 3 0-48 hours",
  "Conv Dose 1 0-24 hours",
  "Conv Dose 2 0-24 hours",
  "Conv Dose 3 0-24 hours",
  "Conv Dose 3 0-48 hours"
)]

combined_gsea_C7 <- data.table::rbindlist(lapply(sel_list_gse_C7, as.data.frame), idcol = TRUE) %>%
  group_by(.id) %>%
  arrange(desc(NES)) %>%
  ungroup()

to_heatmap <- combined_gsea_C7 %>%
  select(.id, ID, NES) %>%
  filter(NES > 3 | NES < -3, !grepl("TBA", ID)) %>%
  tidyr::pivot_wider(values_from = NES, names_from = ID) %>%
  mutate(across(where(anyNA), ~ tidyr::replace_na(., 0))) %>%
  t()
colnames(to_heatmap) <- to_heatmap[1, ]
to_heatmap <- to_heatmap[-1, ]

col_order <- c(
  "Naïve Dose 1 0-24 hours",
  "Conv Dose 1 0-24 hours",
  "Naïve Dose 2 0-24 hours",
  "Conv Dose 2 0-24 hours",
  "Naïve Dose 3 0-24 hours",
  "Conv Dose 3 0-24 hours",
  "Naïve Dose 3 0-48 hours",
  "Conv Dose 3 0-48 hours"
)

to_heatmap <- to_heatmap[, col_order]

C7_heatmap <- matrix(as.numeric(to_heatmap), ncol = ncol(to_heatmap)) %>%
    pheatmap::pheatmap(
      width = 400,
      height = 400,
      cellwidth = 20,
      cellheight = 5,
      cluster_rows = TRUE,
      cluster_cols = FALSE,
      clustering_method = "ward.D2",
      angle_col = 45,
      border_color = "black",
      cutree_cols = 4,
      color = my_palette,
      labels_col = colnames(to_heatmap),
      labels_row = rownames(to_heatmap),
      treeheight_col = 1,
      fontsize_row = 6,
      fontsize_col = 8,
      gaps_col = c(2, 4, 6)
    )
```

```{r, echo=FALSE, results='hide'}
# save C7 heatmap
   ggsave(
      filename = paste0("../", figures.dir,"C7_heatmap.pdf"),
      height = 8, width = 20,
      plot = C7_heatmap)
```

#### Venn Diagram of DEGs

```{r, venn_diagram}
extract_degs_names <- function(x) {
  subsetted <- results_list_plot[[x]] %>%
    tibble::rownames_to_column("gene_symbol") %>%
    filter(log2FoldChange > 1 & padj < 0.05) %>%
    select(gene_symbol, log2FoldChange)
  return(subsetted)
}

degs_names <- lapply(names(results_list_plot), extract_degs_names)
names(degs_names) <- names(results_list_plot)
to_venn <- list(
  conv = degs_names[["Conv 0-24 hours"]][[1]],
  naive = degs_names[["Naïve 0-24 hours"]][[1]]
)

degs_names <- lapply(names(results_list_plot), extract_degs_names)
names(degs_names) <- names(results_list_plot)
to_venn_doses <- list(
  dose1 = degs_names[["Dose 1 0-24 hours"]][[1]],
  dose2 = degs_names[["Dose 2 0-24 hours"]][[1]],
  dose3 = degs_names[["Dose 3 0-24 hours"]][[1]]
)
to_venn_group <- list(
  conv = degs_names[["Conv 0-24 hours"]][[1]],
  naive = degs_names[["Naïve 0-24 hours"]][[1]]
)

ggVennDiagram(to_venn, show_intersect = TRUE)
```

```{r, echo=FALSE, results='hide'}
ggVennDiagram(to_venn_group, show_intersect = TRUE)
# save venn diagram
   ggsave(
      filename = paste0("../", figures.dir,"ggvenndiagram_dose.pdf"),
      height = 7, width = 7)

ggVennDiagram(to_venn_doses, show_intersect = TRUE)
# save venn diagram
   ggsave(
      filename = paste0("../", figures.dir,"ggvenmdiagram_dose.pdf"),
      height = 7, width = 7)
```

## Overrepresentation of gene sets

```{r degs_overrepresentation}
union_doses <- Reduce(base::intersect,to_venn_doses)
union_group <- Reduce(base::intersect,to_venn_group)
unique_conv <- base::setdiff(to_venn_group[[1]],
                              to_venn_group[[2]])
unique_dose2 <- base::setdiff(to_venn_doses[[2]],
                              to_venn_doses[[3]])

list_ema <- list("union_doses" = union_doses, 
                 "union_group" = union_group, 
                 "unique_conv" = unique_conv, 
                 "unique_dose2" = unique_dose2)
enrich_plot_save <- function(x) {
  name_plot <- x
  x <- list_ema[[x]]
  go_enrich <- enrichGO(gene = x, 
           OrgDb = organism,
           keyType = 'SYMBOL',
           ont = "BP",
           pvalueCutoff = 0.05,
           qvalueCutoff = 0.10)

go_enrich_doses <- enricher(gene = unique_dose2,
                     OrgDb = organism, 
                      keyType = 'SYMBOL',
                     TERM2GENE = gmt,
                     pvalueCutoff = 0.05, 
                      qvalueCutoff = 0.10)

  edox2 <- pairwise_termsim(go_enrich)

  emapplot(edox2, layout = "kk", showCategory = 10) +
    scale_color_viridis_c() +
    labs(color = "p adjusted", size = "# genes")
  
  ggsave(
      filename = paste0("../", figures.dir, name_plot,"cmeplot.pdf"),
      height = 7, width = 7)
}

lapply(names(list_ema), enrich_plot_save)

```

## Vaccine detection in RNA-seq

```{r}
tpm_gene_1 <- data.table::fread("../data/processed_data_nfcore/salmon/salmon.merged.gene_tpm.tsv",  header = TRUE, sep = "\t") %>% 
  as.data.frame()

tpm_gene_3 <- data.table::fread("../data/processed_data_nfcore/batch_3/salmon.merged.gene_tpm.tsv",  header = TRUE, sep = "\t") %>% 
  as.data.frame()

tpm_gene <- merge(tpm_gene_1, tpm_gene_3, by=c("gene_id"))

colnames(tpm_gene) <- plyr::mapvalues(sub("_L001|_L002","",colnames(tpm_gene)), from = sampleTable$sample_ID, to = paste(sampleTable$subject_ID,sampleTable$dose, sampleTable$group, sep = "_")) 

names_col <- colnames(tpm_gene)[3:101]
vaccine_tpm <- tpm_gene[64622:64625,]

vaccine_plot <- tidyr::pivot_longer(vaccine_tpm, cols = names_col) %>%
  tidyr::separate(name,into = c("subject_ID","time","dose", "group"),sep = "_", remove = TRUE) %>%
  mutate(time_dose = paste(time,dose, sep = "_"),
         time_dose = factor(time_dose, levels = c("0_dose1", "24_dose1", 
                                        "0_dose2", "24_dose2", 
                                        "0_dose3", "24_dose3", "48_dose3")),
         group = case_when(group == "pos" ~ "Conv",
                           group == "neg" ~ "Naïve"),
         group = factor(group, levels = c("Conv", "Naïve"))) %>%
  na.omit()

vaccine_plot %>%
  filter(grepl("pfizer", gene_id)) %>%
  ggplot(aes(x = time_dose, y = value,group = subject_ID, fill = group)) +
  geom_line() +
  geom_point(shape = 21, size = 5, color = "black", stroke = 1.2) +
  scale_fill_manual(values = c("#A9CAF6","#A9A9A9")) +
  scale_y_log10(limits=c(0.001, 1000), breaks=c(0.01, 0.1, 1, 10, 100))+
  labs(y = "Transcripts per million (log)", x = "") +
  ggprism::theme_prism() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  facet_wrap(~ group, ncol = 4)

 ggsave(
      filename = paste0("../", figures.dir,"pfizer_detection_rna-seq.pdf"),
      height = 5, width = 7)
```



