---
title: "Orebro RNA-seq analysis"
author: "Rodrigo Arcoverde Cerveira & Gustav Joas"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output: html_document
abstract: Örebro study RNA-seq data analysis.
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/lab_book/")})
editor_options: 
  chunk_output_type: inline
---

```{r global.options, include=FALSE}
# Set up figures and chunks messages

knitr::opts_knit$set(echo = TRUE,
                     root.dir = getwd(),
                     fig.width = 6, fig.height = 5,
                     fig.align = "center",
                     out.width = 768,
                     fig.pos = "H",
                     warning = FALSE, 
                     message = FALSE)
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.width = 6, fig.height = 5,
                     fig.align = "center",
                     out.width = 768,
                     fig.pos = "H")

result.dir <- paste0("results/",Sys.Date(),"/")
figures.dir <- paste0("results/",Sys.Date(),"/", "figures/")
# Create result.dir with date in if not existent
ifelse(isFALSE(dir.exists(paste0("../",result.dir))), dir.create(paste0("../",result.dir),recursive = TRUE),"Result directory for today exists already!")
ifelse(isFALSE(dir.exists(paste0("../",figures.dir))), dir.create(paste0("../",figures.dir),recursive = TRUE),"Result directory for today exists already!")

options(stringsAsFactors = FALSE) 
```

## Loading libraries

```{r, message=FALSE}
library(data.table)
library(kableExtra)
library(forcats)
library(caret)
library(C50)
library(mgsub)
library(DESeq2) 
library(pheatmap) 
library(ggplot2) 
library(ggrepel) 
library(RColorBrewer) 
library(limma) 
library(edgeR) 
library(enrichR) 
library(gridExtra) 
library(stringr)
library(ggVennDiagram)
library(dplyr) 
library(ensembldb)
library(EnsDb.Hsapiens.v86)
library(factoextra)
library(rafalib)
library(plotly)
library(tibble)
library(devtools)
library(clusterProfiler)
library(enrichplot)
library(pathview)
```


## Load and wrangle raw counts

```{r}
# Load 1st batch raw counts
counts_raw_1 <- data.table::fread("../data/proces-data_core-facility/subreadCounts_hg38ens_minus_frag.txt", header = TRUE, sep = "\t") %>%  as.data.frame()

# Load 2nd batch raw counts
counts_raw_2 <- data.table::fread("../data/proces-data_core-facility/3rd_subreadCounts_hg38ens_minus_frag.txt", header = TRUE, sep = "\t") %>% as.data.frame()

# Join raw count data frames
counts_raw <- left_join(counts_raw_1, counts_raw_2, by = "Geneid")

# Rename genes based on Emsembl annotation
edb <- EnsDb.Hsapiens.v86
edb_genes <- genes(edb) %>% as.data.frame()
  
# Remove unnecessary columns and convert 'Geneid' to gene names
counts_raw <- counts_raw %>%
  dplyr::select(-contains(c("Chr", "Start", "End", "Strand", "Length"))) %>%
  dplyr::mutate(Geneid = plyr::mapvalues(x = counts_raw$Geneid, 
                                         from = edb_genes$gene_id, 
                                         to = edb_genes$symbol, 
                                         warn_missing = FALSE))
  
# Aggregate and sum up genes with same gene symbol (basically non-coding RNAs)
counts_raw <- aggregate(counts_raw[,-1], list(Geneid=counts_raw[,1]), FUN = sum)

# Change row names to gene names
rownames(counts_raw) <- counts_raw$Geneid
counts_raw <- dplyr::select(counts_raw, -c(Geneid))

```

## Load metadata and match with raw counts
```{r}
# Load sample metadata
sampleTable <- read.table("../data/metadata/sample_metadata.csv", sep = ",", header = TRUE) %>% 
  dplyr::select(-c(age, sex))

# Load patient metadata
patient_metadata <- read.table("../data/metadata/patient_metadata.csv", sep = ",", header = TRUE)

# Join sample and patient metadata
sampleTable <- sampleTable %>% left_join(patient_metadata, by = "subject_ID")

# Change visits to dose and time point...
sampleTable <- sampleTable %>%
 mutate(dose  = as.factor(plyr::mapvalues(visit, 
                                          from = c(1, 2, 4, 5, 10, 11, 12),
                                          to   = c("0_dose1", "24_dose1", "0_dose2", "24_dose2", "0_dose3", "24_dose3", "48_dose3"))),
         group = as.factor(group)) %>% 
#...(sex) 0 and 1 to male and female...
  mutate(sex  = as.factor(plyr::mapvalues(sex, 
                                          from = c(0, 1),
                                          to   = c("male", "female"))),
         group = as.factor(sex)) %>% 
#...(prior covid) 0 and 1 to neg and pos
  mutate(prior_covid19  = as.factor(plyr::mapvalues(prior_covid19, 
                                          from = c(0, 1),
                                          to   = c("neg", "pos"))),
         group = as.factor(prior_covid19))

# Remove 24dose_2 from subject 21 (was missing 0dose_2)
sampleTable<- subset(sampleTable, sample_ID!="P22761_1037_S37")

# Remove one sample from subject 7 (sample replicate)
sampleTable<- subset(sampleTable, sample_ID!="P26208_1060_S62")

# Remove corresponding samples from count table
counts_raw <- subset(counts_raw, select = -c(P26208_1060_S62, P22761_1037_S37))

# Match count data with sample table
colnames(counts_raw) <- sampleTable$sample_ID
counts_raw <- counts_raw[, pmatch(sampleTable$sample_ID, colnames(counts_raw))]
all(rownames(sampleTable$sample_ID) == colnames(counts_raw))
```

## Basic Quality Control

### Inspect raw data

Plotting raw counts for first visual inspection.
Boxplot, the median values are zero across all samples. 
Histogram, huge peak of zeros. This data set would benefit from a low count filtering.
```{r}
# Visualize distribution of raw counts w/ boxplot and density plot
{
  pdf(paste0("../", figures.dir,"raw_counts_QC.pdf"), width = 10,  height = 8, compress = TRUE )
  rafalib::mypar(1,2,mar=c(10,3,3,2))
  boxplot(log2(as.matrix(counts_raw)+1),ylab=expression('Log'[2]~'Read counts'),las=2,main="Raw data")
  hist(log2(as.matrix(counts_raw)+1),ylab="",las=2,main="Raw data")
  par(mfrow=c(1,1))
  dev.off()
}

boxplot(log2(as.matrix(counts_raw)+1),ylab=expression('Log'[2]~'Read counts'),las=2,main="Raw data")
hist(log2(as.matrix(counts_raw)+1),ylab="",las=2,main="Raw data")
```

### Filter data

Plot number of genes detected across samples. All samples are more or less close to average so we don't need to discard any samples.

```{r}
{
  pdf(paste0("../", figures.dir,"number_detected_genes.pdf"), width = 10,  height = 8, compress = TRUE )
  par(mar=c(10,4,3,4))
  barplot(colSums(counts_raw>3),ylab="Number of detected genes",las=2) +
  abline(h=median(colSums(counts_raw>3)))
  dev.off()
}
barplot(colSums(counts_raw>3),ylab="Number of detected genes",las=2)+
abline(h=median(colSums(counts_raw>3)))
 
```

Removing reads with the log2 of the counts per million (cpm) lower than 1 (45896 genes).

```{r}

meanLog2CPM <- rowMeans(log2(cpm(counts_raw) + 1)) 
counts_filtered <- counts_raw[meanLog2CPM > 1, ] 

```

Plot genes detection rate across samples, comparing raw and filtered counts

```{r}
{
  pdf(paste0("../", figures.dir,"detection_rate_raw_filtered.pdf"), width = 10,  height = 8, compress = TRUE )
  par(mar=c(10,4,3,4))
  hist(rowSums(counts_raw>3))
  hist(rowSums(counts_filtered>3))
  par(mfrow=c(1,1))
  dev.off()
}

hist(rowSums(counts_raw>3))
hist(rowSums(counts_filtered>3))
 
```

Plot distribution of the filtered counts

```{r}
{
  pdf(paste0("../", figures.dir,"filtered_counts_distribution.pdf"), width = 10,  height = 8, compress = TRUE )
  rafalib::mypar(1,2,mar=c(10,3,3,2))
  boxplot(log2(as.matrix(counts_filtered)+1),ylab=expression('Log'[2]~'Read counts'),las=2,main="Filtered data")
  hist(log2(as.matrix(counts_filtered)+1),ylab="",las=2,main="Filtered data")
  par(mfrow=c(1,1))
  dev.off()
}

boxplot(log2(as.matrix(counts_filtered)+1),ylab=expression('Log'[2]~'Read counts'),las=2,main="Filtered data")
hist(log2(as.matrix(counts_filtered)+1),ylab="",las=2,main="Filtered data")
  
```

### DESeq object creation, Normalization and Quality Control

Generating the DESeq dataset.

```{r}
# Prepare for DESeq
# Create new variables and factorize needed variables
sampleTable <- sampleTable %>%
  mutate(conditions = paste0(sampleTable$group, "_", sampleTable$dose),
         time = gsub('.{6}$', '', conditions),
         conditions = factor(conditions),
         time = factor(time),
         sex = factor(sex),
         dose = factor(dose),
         group = factor(group))

# create 3 DESeq objects comparing against conditions, time and dose
dds1 <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              colData = as.data.frame(sampleTable),
                              design = ~  sex + age + conditions)

dds2 <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              colData = as.data.frame(sampleTable),
                              design = ~  sex + age + time)

dds3 <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              colData = as.data.frame(sampleTable),
                              design = ~  sex + age + dose)

# Normalize with rlog
counts_rlog_normalized <- rlog(dds1, blind = TRUE)

# Normalize with variance stabilizing transformation for later PCA and heatmap
counts_vst_normalized <- vst(dds1, blind = TRUE)
```

#### Plot distribution of data after normalization

```{r}
vst_matrix <- assay(counts_vst_normalized) 
rlog_matrix <- assay(counts_rlog_normalized)
hist(vst_matrix)
hist(rlog_matrix)
boxplot(vst_matrix ,ylab=expression('Log'[2]~'Read counts'),las=2,main="VST")
boxplot(rlog_matrix ,ylab=expression('Log'[2]~'Read counts'),las=2,main="rlog")

```

### Heatmap

```{r}
# Sample heatmap
sampleDist <- cor(vst_matrix, method = "spearman") 

Metadata <- data.frame(sampleTable$group, sampleTable$dose)
names(Metadata) <- c("Group","Dose")
rownames(Metadata) <- sampleTable$sample_ID

# Plot heatmap
colors<-colorRampPalette(rev(brewer.pal(n=7,name="RdBu")))(255)

{
  pdf(paste0("../", figures.dir,"heatmap.pdf"), width = 10,  height = 8, compress = TRUE )
 Heatmap <-  pheatmap(sampleDist, 
           color = colors,
           clustering_distance_rows = as.dist(1 - sampleDist),
           clustering_distance_cols = as.dist(1 - sampleDist), 
           show_rownames = F,
           show_colnames = F,
           clustering_method = "ward.D2",
           annotation_col = Metadata)
  par(mfrow=c(1,1))
  dev.off()
}

print(Heatmap)
```

### Principal Component Analyais (PCA)

Dimensionality reduction for evaluating outliers and global sample clusters for both quality control but also for primary exploratory analysis.

```{r}
# PCA plot
pcaRes <- prcomp(t(assay(counts_vst_normalized)))
varExp <- round(summary(pcaRes)[[1]],digits = 2)

pcaDF <- data.frame(
  PC1 = pcaRes$x[, 1],
  PC2 = pcaRes$x[, 2],
  Group = sampleTable$group,
  Sample = sampleTable$sample_ID,
  Dose = sampleTable$dose)

{ 
  pdf(paste0("../", figures.dir,"PCA.pdf"), width = 10,  height = 8, compress = TRUE )
  pcaPlot <- ggplot( pcaDF, mapping = aes(x = PC1, y = PC2, color = Group, label = Sample)) + 
  geom_point(aes(shape = Group), size = 3) +
  labs(x = paste0("PC1 (", varExp[1], " %)"), y = paste0("PC2 (", varExp[2], " %)")) + 
  theme(axis.text = element_text(size = 12), legend.text = element_text(size = 10)) +
  scale_color_manual(values = brewer.pal(3, "Accent")) +
  cowplot::theme_cowplot()
  par(mfrow=c(1,1))
  dev.off()
}

print(pcaPlot)

# Plot scree plot 
pca.var <- pcaRes$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
{
  pdf(paste0("../", figures.dir,"Scree_plot.pdf"), width = 10,  height = 8, compress = TRUE )  
  barplot(pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation")
  par(mfrow=c(1,1))
  dev.off()
}
barplot(pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation")

# Explore which variables contribute the most for each PC
var <- get_pca_var(pcaRes)
var$contrib %>%
  as.data.frame() %>%
  arrange(desc(Dim.1)) %>%
  head() %>%
  kableExtra::kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
  
```

## Differential Gene Expression Analysis

```{r}
# Run the DESeq2 analysis
dds1 <-  DESeq(dds1)
dds2 <-  DESeq(dds2)
dds3 <-  DESeq(dds3)

# Contrast
resultsNames(dds1)
resultsNames(dds2)
resultsNames(dds3)

# Create function for makings comparisons with default being conditions and dds1
compare_DEGs <- function(variable1, variable2, condition = "conditions", dds = dds1) {
  as.data.frame(results(dds, contrast = c(condition, variable1, variable2)))
} 

# Baselines DEGs
base_pos_neg_d1 <- compare_DEGs("pos_0_dose1", "neg_0_dose1")
base_pos_neg_d2 <- compare_DEGs("pos_0_dose2", "neg_0_dose2")
base_pos_neg_d3 <- compare_DEGs("pos_0_dose3", "neg_0_dose3")

base_pos_d1_d2 <- compare_DEGs("pos_0_dose2", "pos_0_dose1") 
base_pos_d1_d3 <- compare_DEGs("pos_0_dose3", "pos_0_dose1") 
base_pos_d2_d3 <- compare_DEGs("pos_0_dose2", "pos_0_dose3")

base_neg_d1_d2 <- compare_DEGs("neg_0_dose2", "neg_0_dose1") 
base_neg_d1_d3 <- compare_DEGs("neg_0_dose3", "neg_0_dose1")
base_neg_d2_d3 <- compare_DEGs("neg_0_dose2", "neg_0_dose3")

# 0-24h and 0-48 (3rd dose) DEGs
neg_d1 <- compare_DEGs("neg_24_dose1", "neg_0_dose1")
neg_d2 <- compare_DEGs("neg_24_dose2", "neg_0_dose2")
neg_d3_24 <- compare_DEGs("neg_24_dose3", "neg_0_dose3")
neg_d3_48 <- compare_DEGs("neg_48_dose3", "neg_0_dose3")

pos_d1 <- compare_DEGs("pos_24_dose1", "pos_0_dose1")
pos_d2 <- compare_DEGs("pos_24_dose2", "pos_0_dose2")
pos_d3_24 <- compare_DEGs("pos_24_dose3", "pos_0_dose3")
pos_d3_48 <- compare_DEGs("pos_48_dose3", "pos_0_dose3")

# Results dds2
pos <- compare_DEGs("pos_24", "pos_0", condition = "time", dds = dds2)
neg <- compare_DEGs("neg_24", "neg_0", condition = "time", dds = dds2)

# Results dds3
d1 <- compare_DEGs("24_dose1", "0_dose1", condition = "dose", dds = dds3)
d2 <- compare_DEGs("24_dose2", "0_dose2", condition = "dose", dds = dds3)
d3_24 <- compare_DEGs("24_dose3", "0_dose3", condition = "dose", dds = dds3)
d3_48 <- compare_DEGs("48_dose3", "0_dose3", condition = "dose", dds = dds3)
```


### Volcano plots

Volcano plot for data exploratory analysis, filtering for significant values with False Discovery Rate < 0.05 and log fold change greater than 1.

```{r}
# check only the negative fold change to see if there's a different pahtway of negative fold changes

results_list <- list(base_pos_neg_d1,
                    base_pos_neg_d2,
                    base_pos_neg_d3, 
                    base_pos_d1_d2, 
                    base_pos_d1_d3, 
                    base_pos_d2_d3, 
                    base_neg_d1_d2, 
                    base_neg_d1_d3, 
                    base_neg_d2_d3, 
                    neg_d1,
                    neg_d2, 
                    neg_d3_24, 
                    neg_d3_48, 
                    pos_d1, 
                    pos_d2, 
                    pos_d3_24, 
                    pos_d3_48, 
                    pos, 
                    neg, 
                    d1, 
                    d2, 
                    d3_24, 
                    d3_48)
                     
names(results_list) <- c("Baseline Dose 1 Conv vs Naïve", 
                         "Baseline Dose 2 Conv vs Naïve", 
                         "Baseline Dose 3 Conv vs Naïve", 
                         "Baseline Conv Dose 1 and 2", 
                         "Baseline Conv Dose 1 and 3", 
                         "Baseline Conv Dose 2 and 3", 
                         "Baseline Naïve Dose 1 and 2", 
                         "Baseline Naïve Dose 1 and 3", 
                         "Baseline Naïve Dose 2 and 3",
                         "Naïve Dose 1 0-24 hours", 
                         "Naïve Dose 2 0-24 hours", 
                         "Naïve Dose 3 0-24 hours", 
                         "Naïve Dose 3 0-48 hours", 
                         "Conv Dose 1 0-24 hours", 
                         "Conv Dose 2 0-24 hours", 
                         "Conv Dose 3 0-24 hours", 
                         "Conv Dose 3 0-48 hours", 
                         "Conv 0-24 hours",
                         "Naïve 0-24 hours", 
                         "Dose 1 0-24 hours", 
                         "Dose 2 0-24 hours",
                         "Dose 3 0-24 hours",
                         "Dose 3 0-48 hours") 

    results_list_plot <- lapply(results_list, function(x) {
  x <- x %>% mutate(test_padj = case_when(padj < 0.05 & log2FoldChange >= 1 ~ "Up regulated",
                                     padj < 0.05 & log2FoldChange <= -1 ~ "Down regulated",
                                     TRUE ~ "Not significant"),
                    test_pvalue = case_when(pvalue < 0.05 & log2FoldChange >= 1 ~ "Up regulated",
                                     pvalue < 0.05 & log2FoldChange <= -1 ~ "Down regulated",
                                     TRUE ~ "Not significant"),
                    )
  })
    
# function to plot and save volcano plot
volcano_plot <-  function(x){
  subsetted <- subset(results_list_plot[[x]] %>% tibble::rownames_to_column("gene_symbol"), abs(log2FoldChange) >= 1 & pvalue < 0.05)
  # count up regulated DEGs
  deg_number_up <- subset(results_list_plot[[x]] %>% tibble::rownames_to_column("gene_symbol"), log2FoldChange >= 1 & pvalue < 0.05) %>%
    nrow()
  # count down regulated DEGs
  deg_number_down <- subset(results_list_plot[[x]] %>% tibble::rownames_to_column("gene_symbol"), log2FoldChange <= -1 & pvalue < 0.05) %>%
    nrow()
  
  results_list_plot[[x]] %>% 
  ggplot(aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(colour = test_pvalue), size = 3, alpha = 0.3) +
  scale_color_manual(values = c("Down regulated"="blue", "Not significant" ="grey80", "Up regulated"="red")) +
  xlab(expression("Fold Change (Log"[2]*")")) +
  ylab(expression("-Log"[10]*"(p-value)")) +
  labs(x = NULL, y = NULL) +
  geom_vline(xintercept = c(-1), linetype = "dotted", size = 1) + 
  geom_vline(xintercept = c(1), linetype = "dotted", size = 1) + 
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", size = 1) +
 # geom_label_repel(data = subsetted, aes(log2FoldChange,-log10(pvalue),label= gene_symbol), max.overlaps = 10) +
 xlim(-2,6) +
 ylim(0,20) +
  ggtitle(x) +  
  annotate(geom='text', colour = "red", size = 10, x= 4, y=10, hjust=0, label=paste0(deg_number_up)) +
  annotate(geom='text', colour = "blue", size = 10, x= -2, y= 10, hjust=0, label=paste0(deg_number_down)) +
 ggprism::theme_prism() +
  theme(axis.ticks = element_line(size = .5),legend.position = 'none' ) 
  }

volcano_plots <- lapply(names(results_list_plot), volcano_plot)
names(volcano_plots) <- names(results_list_plot)
volcano_plots

lapply(names(volcano_plots), function(x) ggsave(filename = paste0("../", figures.dir, x,"_volcano_plot.png"), width = 7, height = 7,
                                         plot = volcano_plots[[x]]))
```

### Enrichment analysis

Selecting two different databases for search for enrichment analysis. The databases selected were "GO_Biological_Process_2021" and "KEGG_2021_Human". Testing also the Blood Transcriptome modules presented in [Li et al. 2014](10.1038/ni.2789) using some GSEA package.

```{r}
#Enrichment analysis
listEnrichrDbs()

enrichment_list_up <- lapply(results_list_plot, function(x) {
  enrichr(
  genes = rownames(x[x$test_padj == "Up regulated", ]),
  databases = c("GO_Biological_Process_2021", "KEGG_2021_Human","GO_Molecular_Function_2021"))})


enrichment_list_down <- lapply(results_list_plot, function(x) {
  enrichr(
  genes = rownames(x[x$test_padj == "Down regulated", ]),
  databases = c("GO_Biological_Process_2021", "KEGG_2021_Human","GO_Molecular_Function_2021"))})

gmt <- read.gmt("../data/blood_transcription_module/BTM_for_GSEA_20131008.gmt")

```

#### Visualize enriched pathways

```{r echo=TRUE, fig.width = 15, fig.height = 8}
# Remove items without significant results from list
enrichment_list_up[sapply(enrichment_list_up, is.null)] <- NULL
enrichment_list_down[sapply(enrichment_list_down, is.null)] <- NULL

# Optional: remove all comparisons except naive vs conv at each timepoint
  # enrichment_list_up <- enrichment_list_up[3:9]
  # enrichment_list_down <- enrichment_list_down[3:6]

# Plot over-representation analysis
plot_enrichment <- function(enrichment_list){
 plots_enrich <- list()
for(pathway in 1:length(enrichment_list[[1]])){
  
  rbinded_enrich_list <- lapply(enrichment_list, function(x)x[[pathway]])
  print(names(enrichment_list[[1]])[[pathway]])
  rbinded_enrich_df <- data.table::rbindlist(l = rbinded_enrich_list, idcol = TRUE, fill = TRUE)
  
  plots_enrich[[pathway]] <- rbinded_enrich_df %>%
  group_by(.id) %>%
  arrange(Adjusted.P.value) %>%
  slice(1:10) 
  }
plots <- lapply(plots_enrich, function(x){
  x %>%
  ggplot(aes(x = Term, y = -log10(Adjusted.P.value))) +
    geom_bar(stat = "identity", width = 0.05) + geom_point(size = 3) +
    theme_minimal() +
    theme(text = element_text(size = 10),
          plot.title = element_text(hjust = (5 / nchar(results_list)) * 2), 
          plot.margin = margin(t = 5, r = 50, b = 5, unit = "pt"), 
          axis.text.y = element_text(size = 8)) +
    coord_flip() +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray") + 
    theme_bw() +
    labs(x = NULL, y = expression("Adjusted P value, Log"[10]*"")) +
  facet_wrap(~.id, ncol = 4)
  })
  return(plots)
}

# Plotting over-representation analyssi
enrichment_plot_up <- plot_enrichment(enrichment_list_up)
enrichment_plot_down <- plot_enrichment(enrichment_list_down)

names(enrichment_plot_up) <- c("GO_Biological_Process_2021_up",
                            "KEGG_2021_Human_up",
                            "GO_Molecular_Function_2021_up") 

names(enrichment_plot_down) <- c("GO_Biological_Process_2021_down",
                            "KEGG_2021_Human_down",
                            "GO_Molecular_Function_2021_down") 
print(enrichment_plot_up)
print(enrichment_plot_down)
 

# function for saving op plots
save_op_plots <- function(ea_plot_list){
  lapply(names(ea_plot_list), function(x) 
     ggsave(filename = paste0("../", figures.dir, x,"_EA.png"), 
            height = 24, width = 16, 
            plot = ea_plot_list[[x]]))
}

# saving enrichment plots to figure folder
save_op_plots(enrichment_plot_up)
save_op_plots(enrichment_plot_down)

```

### Gene Set Enrichment Analysis with ClusterProfiler

#### Annotations

```{r, message=F, warning=F}
# SET THE DESIRED ORGANISM HERE
organism <- "org.Hs.eg.db"
# BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
```

#### Prepare Input

```{r}
process_to_gsea <- function(x){
  original_gene_list <- x$log2FoldChange
  names(original_gene_list) <- rownames(x)
  gene_list<-na.omit(original_gene_list)
  gene_list <- sort(gene_list, decreasing = TRUE)
}

ls_processed <- lapply(results_list, process_to_gsea)

```

#### Gene Set Enrichment

```{r}
list_gse <- lapply(ls_processed, function(x) GSEA(geneList=x, 
            TERM2GENE = gmt,
             minGSSize = 3, 
             maxGSSize = 800, 
            nPermSimple = 10000,
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             pAdjustMethod = "fdr"))


```


#### Ridgeplot

Helpful to interpret up/down-regulated pathways.

```{r fig.width=18, fig.height=12}
# try also to plot as a heatmap for all the group comparisons

ridgeplotting <- function(x){
  list_gse[[x]] %>%
    ridgeplot(fill = "enrichmentScore") +
    viridis::scale_fill_viridis(option = "turbo") +
    ggtitle(x) +
    cowplot::theme_cowplot(font_size = 6, 
                         line_size = 1.5, 
                         rel_small = 1.6, 
                         rel_tiny = 1,
                         font_family = "Arial",
                         rel_large = 4) +
  theme(legend.position = 'right',
        legend.direction = 'vertical',
        legend.key.size = unit(1, 'cm'),
        legend.title = element_text(size = 16, face = "bold", family = "Arial"),
        legend.text = element_text(size = 14),
        axis.title = element_text(size = 35), 
        axis.text.x  = element_text(size = 16),
        axis.text.y =  element_text(size = 16)) +
  scale_y_discrete()
}

ridgeplots <- lapply(names(list_gse), ridgeplotting)
names(ridgeplots) <- names(list_gse)
print(ridgeplots)

lapply(names(ridgeplots), function(x) ggsave(filename = paste0("../", figures.dir, x,"_BTM_plot.png"), 
                                         plot = ridgeplots[[x]], width = 16, height = 12))
```


#### GSEA Plot  
Traditional method for visualizing GSEA result.  
  
```{r fig.height=6, fig.width=8}
# Use the `Gene Set` param for the index in the title, and as the value for geneSetId
#gseaplot(gse, by = "all", title = gse$Description[55], geneSetID = 55)
#head(gse, 100)

```

### Visualization of DEGS
#### Venn Diagram of DEGs

```{r venn_diagram}
extract_degs_names <-  function(x){
  subsetted <- results_list_plot[[x]] %>% tibble::rownames_to_column("gene_symbol") %>%
    filter(log2FoldChange > 1 & padj < 0.05) %>%
    select(gene_symbol, log2FoldChange)
  return(subsetted)
}

degs_names <- lapply(names(results_list_plot), extract_degs_names)
names(degs_names) <- names(results_list_plot)
to_venn <- list(conv = degs_names[["Conv 0-24 hours"]][[1]], 
                naive = degs_names[["Naïve 0-24 hours"]][[1]])

ggVennDiagram(to_venn, show_intersect = TRUE)
```



### Session info

```{r}
sessionInfo()
```
